---
title: "Homework 6: Markov chain Monte Carlo"
author:
  - Bernhard Buss
  - Dario Franzen
  - Aurimas Greicius
  - Timothy Lauren Nijhuis
  - Carlo Rossetto
  - Yuliyan Slavchev
  - Görkem Kadir Solun
format: pdf
editor: visual
---

```{r}
set.seed(42)
```

## Problem 15: Markov chain Monte Carlo estimation of the mean

### Part a: Proof that $\overline Y$ is an unbiased estimator of the mean

Given that $$\overline Y = \frac{1}{N}\sum^{N}_{i=1}Y_{n}$$ $$E\left[\overline Y\right]=E\left[\frac{1}{N}\sum^{N}_{n=1}Y_{n}\right]=\frac{1}{N}\sum^{N}_{n=1}E\left[Y_{n}\right]=$$ $$\frac{1}{N}\sum^{N}_{n=1}\mu=\frac{1}{N}N\mu=\mu$$

### Part b: Approximation of the variance of $\overline Y$

Knowing that $$Var(\overline Y)=Var\left(\frac{1}{N}\sum^{N}_{n=1}Y_{n}\right)=\frac{Var\left(\sum^{N}_{n=1}Y_{n}\right)}{N^2}=$$ When we start from Bienayme's identity an we are multiplying by $\frac{1}{N^2}$: $$\frac{1}{N^2}Var\left(\sum^{N}_{n=1}Y_{n}\right)=\frac{1}{N^2}\sum^{N}_{i=1}\sum^{N}_{j=1}Cov(Y_{i},Y_{j})=$$ $$\frac{1}{N^2}\left(\sum^{N}_{i=1}Var(Y_{i})+2\sum_{1\le i<j\le N}Cov(Y_{i},Y_{j})\right)=$$ $$\frac{1}{N^2}\left(N\sigma^2+2\sum_{k=1}^{N-1}(N-k)Cov(Y_{n},Y_{n+k})\right)$$Knowing that

$$\rho_k = \frac{Cov(Y_n,Y_{n+k})}{\sigma^2} \implies Cov(Y_n,Y_{n+k})=\sigma^2\rho_k$$

So

$$\frac{1}{N^2}\left(N\sigma^2+2\sum_{k=1}^{N-1}(N-k)Cov(Y_{n},Y_{n+k})\right)=\frac{1}{N^2}\left(N\sigma^2+2\sum_{k=1}^{N-1}(N-k)\sigma^2\rho_k\right)=$$

$$\frac{N\sigma^2}{N^2}\left(1+\frac{2}{N}\sum_{k=1}^{N-1}(N-k)\rho_k\right)=\frac{\sigma^2}{N}\left(1+2\left(\frac{1}{N}\sum_{k=1}^{N-1}(N-k)\rho_k\right)\right)$$ Since for large N $$\frac{1}{N}\sum_{k=1}^{N-1}(N-k)\rho_{k}\approx \sum_{k=1}^{\infty}\rho_{k}$$ $$\frac{\sigma^2}{N}\left(1+2\left(\frac{1}{N}\sum_{k=1}^{N-1}(N-k)\rho_k\right)\right)=\frac{\sigma^2}{N}\left(1+2\sum_{k=1}^{\infty}\rho_k\right)$$ Lastly, since $$N_{eff}=\frac{N}{1+2\sum_{k=1}^{\infty}\rho_{k}}$$ $$\sigma^2\frac{\left(1+2\sum_{k=1}^{\infty}\rho_{k}\right)}{N}=\frac{\sigma^2}{N_{eff}}\approx Var\left(\overline Y\right)$$

## Problem 16: Sampling in the Rain Network

### Part a: Joint probability and naive marginalization

$$P(C=T|R=T,S=T,W=T) = \frac{P(C=T,R=T,S=T,W=T)}{P(R=T,S=T,W=T)}$$ The numerator: $$P(C=T,R=T,S=T,W=T) = $$ $$ P(W=T|S=T,R=T)*P(S=T|C=T)*P(R=T|C=T)*P(C=T)=$$ $$0.99*0.10*0.80*0.50=0.0396$$ The denominator: $$P(R=T,S=T,W=T) = $$ $$P(W=T|S=T,R=T)*P(S=T|C=T)*P(R=T|C=T)*P(C=T)+$$ $$P(W=T|S=T,R=T)*P(S=T|C=F)*P(R=T|C=F)*P(C=F)=$$ $$0.99*0.10*0.80*0.50+0.99*0.50*0.20*0.50=0.0396+0.0495=0.0891$$ Substituting into the fraction: $$P(C=T|R=T,S=T,W=T) = \frac{P(C=T,R=T,S=T,W=T)}{P(R=T,S=T,W=T)}=$$ $$ \frac{0.0396}{0.0891}=0.\overline{44}=\frac{4}{9}\approx0.444$$

$$P(C=T|R=F,S=T,W=T) = \frac{P(C=T,R=F,S=T,W=T)}{P(R=F,S=T,W=T)}$$ The numerator: $$P(C=T,R=F,S=T,W=T) = $$ $$ P(W=T|S=T,R=F)*P(S=T|C=T)*P(R=F|C=T)*P(C=T)=$$ $$0.90*0.10*(1-0.80)*0.50=0.009$$ The denominator: $$P(R=F,S=T,W=T) = $$ $$P(W=T|S=T,R=F)*P(S=T|C=T)*P(R=F|C=T)*P(C=T)+$$ $$P(W=T|S=T,R=F)*P(S=T|C=F)*P(R=F|C=F)*P(C=F)=$$ $$0.90*0.10*(1-0.80)*0.50+0.90*0.50*(1-0.20)*0.50=0.009+0.180=0.189$$ Substituting into the fraction: $$P(C=T|R=F,S=T,W=T) = \frac{P(C=T,R=F,S=T,W=T)}{P(R=F,S=T,W=T)}=$$ $$ \frac{0.009}{0.189}=0.047619\approx0.048$$ $$P(R=T|C=T,S=T,W=T) = \frac{P(R=T,C=T,S=T,W=T)}{P(C=T,S=T,W=T)}$$ The numerator: $$P(R=T,C=T,S=T,W=T) = $$ $$ P(W=T|S=T,R=T)*P(S=T|C=T)*P(R=T|C=T)*P(C=T)=$$ $$0.99*0.10*0.80*0.50=0.0396$$ The denominator: $$P(C=T,S=T,W=T) = $$ $$P(W=T|S=T,R=T)*P(S=T|C=T)*P(R=T|C=T)*P(C=T)+$$ $$P(W=T|S=T,R=F)*P(S=T|C=T)*P(R=F|C=T)*P(C=T)=$$ $$0.99*0.10*0.80*0.50+0.90*0.10*0.20*0.50=0.0396+0.009=0.0486$$ Substituting into the fraction: $$P(R=T|C=T,S=T,W=T) = \frac{P(R=T,C=T,S=T,W=T)}{P(C=T,S=T,W=T)}=$$ $$ \frac{0.0396}{0.0486}=0.814815\approx0.815$$ $$P(R=T|C=F,S=T,W=T) = \frac{P(R=T,C=F,S=T,W=T)}{P(C=F,S=T,W=T)}$$ The numerator: $$P(R=T,C=F,S=T,W=T) = $$ $$ P(W=T|S=T,R=T)*P(S=T|C=F)*P(R=T|C=F)*P(C=F)=$$ $$0.99*0.50*0.20*0.50=0.0495$$ The denominator: $$P(C=F,S=T,W=T) = $$ $$P(W=T|S=T,R=T)*P(S=T|C=F)*P(R=T|C=F)*P(C=F)+$$ $$P(W=T|S=T,R=F)*P(S=T|C=F)*P(R=F|C=F)*P(C=F)=$$ $$0.99*0.50*0.20*0.50+0.90*0.50*(1-0.20)*0.50=0.0495+0.1800=0.2295$$ Substituting into the fraction: $$P(R=T|C=F,S=T,W=T) = \frac{P(R=T,C=F,S=T,W=T)}{P(C=F,S=T,W=T)}=$$ $$ \frac{0.0495}{0.2295}=0.215686\approx0.216$$ In summary: $$P(C=T|R=T,S=T,W=T) \approx 0.444$$ $$P(C=T|R=F,S=T,W=T) \approx 0.048$$ $$P(R=T|C=T,S=T,W=T) \approx 0.815$$ $$P(R=T|C=F,S=T,W=T) \approx 0.216$$

### Part b: Gibbs sampler for Bayesian network implementation

```{r}
tot_samples <- 100 

SW_Gibbs_samples <- matrix(0, nrow = 2, ncol = tot_samples) 
rownames(SW_Gibbs_samples) <- c("C", "R")

# initialising C and R with random starting states
current_C <- rbinom(1,1,0.5)   # rbinom(n=1, size=1, prob) flips a weighted coin
current_R <- rbinom(1,1,0.5)

for (n in seq(tot_samples)) {
  
  # sample C
  if (current_R == 1) {
    prob_C <- 0.4444  # P(C=T | R=T, S=T, W=T)
  } else {
    prob_C <- 0.0476  # P(C=T | R=F, S=T, W=T)
  }

  current_C <- rbinom(1, 1, prob_C) 
  
  # sample R
  if (current_C == 1) {
    prob_R <- 0.8148  # P(R=T | C=T, S=T, W=T)
  } else {
    prob_R <- 0.2157  # P(R=T | C=F, S=T, W=T)
  }
  current_R <- rbinom(1, 1, prob_R)
  
  # recording the current values
  SW_Gibbs_samples["C", n] <- current_C
  SW_Gibbs_samples["R", n] <- current_R
}  
```

### Part c: Marginal probability of rain

```{r}
count_rain <- rowSums(SW_Gibbs_samples)["R"]
count_clouds <- rowSums(SW_Gibbs_samples)["C"] # not asked

p_rain <- count_rain / tot_samples
p_clouds <- count_clouds /tot_samples # not asked

print(p_rain)
```

The marginal probability of rain when the sprinkler is on and the grass is wet, $P(R=T|S=T,W=T)$, is `r p_rain`

### Part d: Draw 50,000 samples using the Gibbs sampler

```{r}
Gibbs_sampler <- function(tot_samples = 50000) {

  SW_Gibbs_samples <- matrix(0, nrow = 2, ncol = tot_samples)
  rownames(SW_Gibbs_samples) <- c("C", "R")
  
  # initialising C and R with random starting states
  current_C <- rbinom(1,1,0.5)   # rbinom(n=1, size=1, prob) flips a weighted coin
  current_R <- rbinom(1,1,0.5)
  
  for (n in seq(tot_samples)) {
    
    # sample C
    if (current_R == 1) {
      prob_C <- 0.4444  # P(C=T | R=T, S=T, W=T)
    } else {
      prob_C <- 0.0476  # P(C=T | R=F, S=T, W=T)
    }
  
    current_C <- rbinom(1, 1, prob_C) 
    
    # sample R
    if (current_C == 1) {
      prob_R <- 0.8148  # P(R=T | C=T, S=T, W=T)
    } else {
      prob_R <- 0.2157  # P(R=T | C=F, S=T, W=T)
    }
    current_R <- rbinom(1, 1, prob_R)
    
    # recording the current values
    SW_Gibbs_samples["C", n] <- current_C
    SW_Gibbs_samples["R", n] <- current_R
  } 
  
  return(SW_Gibbs_samples)
}
```

### Part e: Relative frequencies of R=T and C=T and burn-in time

```{r}
tot_samples = 50000

run_1_results <- Gibbs_sampler(tot_samples)
run_2_results <- Gibbs_sampler(tot_samples)
```

```{r}
# correct path or remove entirely :)
setwd("~/My Documents/Carlo/ETH UBas UZH - MSc CBB/636-0702-00L Statistical Models in Computational Biology/Homeworks_GitHub_Linked/hw6")

library(ggplot2)

run_1_c_freq <- cumsum(run_1_results["C", ]) / seq_len(tot_samples)
run_1_r_freq <- cumsum(run_1_results["R", ]) / seq_len(tot_samples)
run_2_c_freq <- cumsum(run_2_results["C", ]) / seq_len(tot_samples)
run_2_r_freq <- cumsum(run_2_results["R", ]) / seq_len(tot_samples)

# Combine the data into a single data frame for ggplot
parte_plot_data <- data.frame(
  t = rep(seq(tot_samples), 4),
  Frequency = c(
    run_1_c_freq,
    run_1_r_freq,
    run_2_c_freq,
    run_2_r_freq
  ),
  Variable = rep(c("C = True", "R = True", "C = True", "R = True"), each = tot_samples),
  Run = rep(c("Run 1", "Run 1", "Run 2", "Run 2"), each = tot_samples)
)

ggplot(parte_plot_data, aes(x = t, y = Frequency, color = Variable, linetype = Run)) +
  geom_line(linewidth = 0.8, alpha = 0.8) +
  theme_minimal() +
  scale_color_manual(values = c("C = True" = "#D55E00", "R = True" = "#0072B2")) +
  labs(
    title = "Frequencies of C=True and R=True over time",
    x = "Iteration (t)",
    y = "Relative Frequency",
    color = "Variable",
    linetype = "Independent Run"
  ) +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, face = "bold")
  )
```

To distinguish more clearly between variables, below are separate plots with different colors per run.

```{r}
ggplot(parte_plot_data, aes(x = t, y = Frequency, color = Run)) +
  geom_line(linewidth = 0.8, alpha = 0.8) +
  theme_minimal() +
  scale_color_manual(values = c("Run 1" = "#009E73", "Run 2" = "#CC79A7")) +
  labs(
    title = "Frequencies of C=True and R=True over time",
    x = "Iteration (t)",
    y = "Relative Frequency",
    color = "Independent Run" # Legend title
  ) +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, face = "bold")
  ) +
  facet_wrap(~ Variable, ncol = 1, scales = "free_y")
```

Based on the trace plots above, a preliminary burn-in of about 20,000 iterations seems reasonable, as there are no fluctuations after this point, and it considers a safety margin.

### Part f: Gelman and Rubin test

```{r}

library(coda)

# transpose so iterations are rows and select the raw R and C counts
combined_mcmc <- mcmc.list(mcmc(t(run_1_results[1:2,])),mcmc(t(run_2_results[1:2,])))

gelman.diag(combined_mcmc)
```

```{r, fig.height=4.8}
gelman.plot(combined_mcmc[, "C", drop = FALSE], auto.layout = FALSE, main = "Gelman-Rubin: C")
```

```{r, fig.height=4.8}
gelman.plot(combined_mcmc[, "R", drop = FALSE], auto.layout = FALSE, main = "Gelman-Rubin: R")
```

The shrink factor drops rapidly to one. For both variables, after about 10,000 iterations the shrink factor is very close to 1, so 10,000 iterations are recommended for warm-up (burn-in). This is lower than the previously proposed 20,000, decreasing the data loss.

### Part g: "Rain and Cloudy" plots

```{r}
acf(run_1_results["C",],type = c("correlation"))
```

```{r}
acf(run_1_results["R",],type = c("correlation")) # lagmax parameter (not included, 30 to 40 by default) is with how many neighbours it calculates the auto-correlation function with
```

After 4 neighbours the autocorrelation function results in a very small value, less than 0.05 (and within the dashed boundary lines), therefore the authors would assume that the 5th sample is independent of the first.

### Part h: Estimation of P(R=T\|S=T,W=T)

Burning the first 30,000 samples and considering only every 5th sample (thinning)

```{r}
run_1_raw_burnt <- run_1_results[1:2,10001:50000]
run_2_raw_burnt <- run_2_results[1:2,10001:50000]

run_1_raw_burnt_thinned <- run_1_raw_burnt[,seq(1,ncol(run_1_raw_burnt),5)]
run_2_raw_burnt_thinned <- run_2_raw_burnt[,seq(1,ncol(run_2_raw_burnt),5)]

freq_run_1 <- rowMeans(run_1_raw_burnt_thinned)
freq_run_2 <- rowMeans(run_2_raw_burnt_thinned)

# taking the mean across runs
parth_R_mean <- mean(c(freq_run_1["R"],freq_run_2["R"]))
parth_C_mean <- mean(c(freq_run_1["C"],freq_run_2["C"]))

print(freq_run_1)
print(freq_run_2)
print(parth_R_mean)
```

### Part i: Analytical computation of P(R=T\|S=T,W=T) and comparison with c and h

$$P(R=T|S=T,W=T) = \frac{P(R=T,S=T,W=T)}{P(S=T,W=T)}$$ The numerator: $$P(R=T,S=T,W=T) = P(W=T|S=T,R=T)*P(S=T|C=T)*P(R=T|C=T)*P(C=T)=$$ $$P(W=T|S=T,R=T)*P(S=T|C=F)*P(R=T|C=F)*P(C=F)+$$ $$0.99*0.10*0.80*0.50+0.99*0.5*0.2*0.5=0.0396+0.0495=0.0891$$ The denominator: $$P(S=T,W=T) = $$ $$P(W=T|S=T,R=T)*P(S=T|C=T)*P(R=T|C=T)*P(C=T)+$$ $$P(W=T|S=T,R=F)*P(S=T|C=T)*P(R=F|C=T)*P(C=T)+$$ $$P(W=T|S=T,R=T)*P(S=T|C=F)*P(R=T|C=F)*P(C=F)+$$ $$P(W=T|S=T,R=F)*P(S=T|C=F)*P(R=F|C=F)*P(C=F)=$$ $$0.99*0.10*0.80*0.50+0.90*0.10*0.20*0.50+$$ $$0.99*0.5*(1-0.8)*0.5+0.9*0.5*(1-0.2)*0.5=$$ $$ 0.0396+0.009+0.0495+0.18=0.2781$$ Substituting into the fraction: $$P(R=T|S=T,W=T) = \frac{P(R=T,S=T,W=T)}{P(S=T,W=T)}=$$ $$ \frac{0.0891}{0.2781}=0.320388\approx0.320$$\
In summary, running two runs with 50,000 samples (burning 10,000 and thinning by taking every fifth) resulted in a marginal probability of rain given the grass is wet and the sprinkler is on, P(R=T\|W=T,S=T), closer to the analytic truth. In fact, part h (mean of two runs with 10,000 samples) resulted in `r parth_R_mean`, while part c (with 100 samples) results in `r p_rain`, which is further away from the analytic result of 0.320.

```{r}
comparison_table <- data.frame(
  Method = c(
    "Part c: 100 samples",
    "Part h: 50,000 samples with burn-in and thinning",
    "Part i: analytical result"
  ),
  Probability = c(
    p_rain,
    parth_R_mean,
    0.320388
  )
)

knitr::kable(
  comparison_table,
  digits = 3,
  col.names = c("Method", "P(R=T | S=T, W=T)"),
  caption = "Comparison of estimated and analytical probabilities"
)
```