In this exercise, we will use the R package BiDAG1
to learn Bayesian networks from protein data.
Download the data here. The data consists of the measurements of 11 phosphorylated proteins
and phospholipids derived from primary immune system cells, subjected to both general and specific
molecular interventions. (Hint: read the help files of the package and use default parameters
unless otherwise stated.)

(a) First, run set.seed(42) for reproducibility. Read in the data in R. Report the number of
variables n and the number of observations N. Visualize the data in a pair plot using the
ggpairs function in the R package GGally. Randomly split the data into 80% training data
and 20% test data. Initialize the parameters using the function BiDAG::scoreparameters
with the training data and the Bayesian Gaussian equivalent (BGe) score.
[Note: The BGe score is a fully-decomposable marginal likelihood function P(D | G) for
scoring Bayesian networks. The main underlying assumption is that the data is normally
distributed with N (µ, W^-1). The precision matrix W follows a Wishart prior Wn(T^-1, αw),
where αw > n − 1 is the degrees of freedom and T is the positive definite parametric matrix.
The mean vector µ follows a normal prior N (ν, Wα_µ) with α_µ > 0.]

Run install.packages("BiDAG") in the R console. Then load the package by running library(BiDAG). You
may need to install the missing packages graph, RBGL, and Rgraphviz from Bioconductor.
```{r}
set.seed(42)
data <- read.csv("hw1/2005_sachs_2_cd3cd28icam2_log_std.csv")
n <- ncol(data)
N <- nrow(data)
cat("Number of variables (n):", n, "\n")
cat("Number of observations (N):", N, "\n")

library(GGally)
ggpairs(data)

train_indices <- sample(seq_len(N), size = floor(0.8 * N))
train_data <- data[train_indices, ]
test_data <- data[-train_indices, ]

library(BiDAG)
myScore <- scoreparameters(score = "bge", data = train_data)
print(myScore)
```

(b) Learn a Bayesian network using the BiDAG::iterativeMCMC function. Plot the directed
acyclic graph (DAG). Evaluate the log BGe score of the test data against the estimated DAG
using the function BiDAG::scoreagainstDAG.

```{r}
library(BiDAG)
library(Rgraphviz)
library(graph)

myDAG <- iterativeMCMC(myScore)

g <- as(myDAG$DAG, "graphNEL")
plot(g) 

testScore <- scoreparameters(score = "bge", data = test_data)

test_log_score <- sum(scoreagainstDAG(scorepar = testScore, incidence = myDAG$DAG))

cat("Log BGe score of test data against estimated DAG:", test_log_score, "\n")
```