---
title: "Homework 8: Structured sparsity in genetics"
author:
  - Bernhard Buss
  - Dario Franzen
  - Aurimas Greicius
  - Timothy Lauren Nijhuis
  - Carlo Rossetto
  - Yuliyan Slavchev
  - Görkem Kadir Solun
format: pdf
editor: visual
---

```{r}
set.seed(42)
```

## Problem 19: Exponential family of the negative binomial distribution

### Part a: Proof that the negative binomial distribution belongs to the exponential family and definition of all its components.

### Part b: Expected value manipulation and derivation

$$
Q(\pi,p,z)=Q(\pi)\prod_{k=1}^KQ(p_k)\prod_{i=1}^NQ(z_i),
$$

## Problem 20: Sparsity and metabolic engineering in B. subtilis

### Part a: Data preparation

```{r}
suppressPackageStartupMessages(library("glmnet"))

# loading the data
riboflavin <- read.csv("riboflavin.csv")
dim(riboflavin)
riboflavin
```

```{r}
set.seed(42)

# splitting into training and test set

samples <- sample(c(TRUE,FALSE), nrow(riboflavin), replace = TRUE, prob = c(0.7,0.3))

r_train <- riboflavin[samples,]
r_test <- riboflavin[!samples,]

x_train <- as.matrix(subset(r_train, select = -y))
y_train <- as.numeric(r_train$y) # cv.glmnet expects a list, not a matrix, so no subset function

x_test <- as.matrix(subset(r_test, select = -y))
y_test <- as.numeric(r_test$y)

dim(x_train)
```
### Part b: Model fitting

### Part c: Sparsity and performance

Part b and c are completed together below.

```{r}
model_alphas <- c(1.0, 0.0, 0.0, 0.0, 0.0, 0.0, 0.0, 0.0, 0.5, 0.0, 0.0, 0.0)
res <- array(model_alphas, dim = c(4,3))
colnames(res) <- c("Lasso regression", "Ridge regression", "Elastic net")
rownames(res) <- c("alpha", "optimal lambda", "number of non-zero coefficients", "MSE on test set")
```

```{r}
for (model in colnames(res)) {
  cv_model <- cv.glmnet(x_train,y_train, alpha = res[1, model], nfolds = 10) #could add type.measure = "mse" as well
  cat("Optimal lambda for", model, " is", cv_model$lambda.min,".\n")
  
  # part b, optimal lambda
  res[2, model] <- cv_model$lambda.min
  
  plot(cv_model)

  # part c, non zero coefficient (expect intercept)
  res[3, model] <- cv_model$nzero[cv_model$lambda == cv_model$lambda.min]
  
  # part c, MSE of prediction on test set
  test_pred <- predict(cv_model,newx = x_test, s = "lambda.min")
  res[4,model] <- sum((test_pred - y_test)^2)/length(test_pred)
  
}
```

```{r}
print(res)
```

## Problem 21: Identification of protein-protein interaction networks via the graphical lasso

### Part a: Raw data loading

```{r}

#loading the data
pe <- read.csv("proteinexpression.csv")
pe

```

### Part b: Lasso model fitting

### Part c: Building the interaction matrix

### Part d: Network plotting


