---
title: "Homework 12: Tumor Phylogenetics"
author:
  - Bernhard Buss
  - Dario Franzen
  - Aurimas Greicius
  - Timothy Lauren Nijhuis
  - Carlo Rossetto
  - Yuliyan Slavchev
  - Görkem Kadir Solun
format: 
  pdf:
    keep-tex: true
    header-includes:
      - \usepackage{cancel}
      - \usepackage{xcolor}
---

```{r}
set.seed(42)
```

## Problem 29: Diploid genotype error model

Calculate the following probabilities analogously:

1.  $W = TT$, and $Y = AG$

$$
\begin{aligned}
  P(TT\mid AG)  =&\ \frac{\delta}{2}\frac{\epsilon}{2}\frac{1}{3} &&\text{dropout with sequencing error } \textcolor{red}{A}\cancel{G} \rightarrow TT \\
                &+ \frac{\delta}{2}\frac{\epsilon}{2}\frac{1}{3} &&\text{dropout with sequencing error } \cancel{A}\textcolor{red}{G} \rightarrow TT \\
                =&\ \frac{\delta \epsilon}{6}
\end{aligned}
$$

2.  $W = AT$, and $Y = AG$

$$
\begin{aligned}
  P\ (AT\mid AG)  =&\ (1-\delta)\frac{\epsilon}{2}\frac{1}{3} &&\text{no dropout, sequencing error } A\textcolor{red}{G}\rightarrow AT \\
                  =&\ \frac{\epsilon}{6} - \frac{\delta\epsilon}{6}
\end{aligned}
$$

3.  $W = AG$, and $Y = AG$

$$
\begin{aligned}
  P\ (AG\mid AG)  =&\ (1-\delta)(1-\epsilon) &&\text{no dropout without sequencing error }AG\rightarrow AG \\
                  =&\ 1 - \delta - \epsilon + \delta\epsilon
\end{aligned}
$$

4.  $W = AA$, and $Y = AA$

$$
\begin{aligned}
  P\ (AA\mid AA)  =&\ (1-\delta)(1-\epsilon) &&\text{no dropout without sequencing error } AA\rightarrow AA \\
                  &+ \frac{\delta}{2}(1-\frac{\epsilon}{2}) &&\text{dropout without sequencing error }\cancel{A}A \rightarrow AA\\
                  &+ \frac{\delta}{2}(1-\frac{\epsilon}{2}) &&\text{dropout without sequencing error }A\cancel{A} \rightarrow AA\\
                  =&\ 1 - \epsilon + \frac{\delta\epsilon}{2}
\end{aligned}
$$

5.  $W = AG$, and $Y = AA$

$$
\begin{aligned}
  P\ (AG\mid AA)  =&\ (1-\delta)\frac{\epsilon}{2}\frac{1}{3} &&\text{no dropout, sequencing error } A\textcolor{red}{A}\rightarrow AG \\
                  &+ (1-\delta)\frac{\epsilon}{2}\frac{1}{3} &&\text{no dropout, sequencing error } \textcolor{red}{A}A\rightarrow AG \\
                  =&\ \frac{\epsilon}{3} - \frac{\delta\epsilon}{3}
\end{aligned}
$$

6.  $W = GG$, and $Y = AA$

$$
\begin{aligned}
  P\ (GG\mid AA)  =&\ \frac{\delta}{2}\frac{\epsilon}{2}\frac{1}{3} &&\text{dropout with sequencing error } \textcolor{red}{A}\cancel{A} \rightarrow GG \\
                &+ \frac{\delta}{2}\frac{\epsilon}{2}\frac{1}{3} &&\text{dropout with sequencing error } \cancel{A}\textcolor{red}{A} \rightarrow GG \\
                =&\ \frac{\delta \epsilon}{6}
\end{aligned}
$$

\newpage

## Problem 30: The Beta-Binomial distribution

For a binomial random variable $X \sim \mathrm{Binom}(n,p)$,

$$
\mathbb{E}[X] = np, \qquad \mathrm{Var}(X) = np(1-p).
$$

For a beta-binomial random variable $X \sim \mathrm{BetaBinom}(n,\alpha,\beta)$,

$$
\mathbb{E}[X] = n\frac{\alpha}{\alpha+\beta},
$$

and

$$
\mathrm{Var}(X)
= n\frac{\alpha\beta}{(\alpha+\beta)^2}
  \frac{\alpha+\beta+n}{\alpha+\beta+1}.
$$

In all four requested cases, the distribution is symmetric around $500$, so the mean is $500$. The variances differ strongly because the beta binomial contains extra variability from the random success probability.

```{r}
n_trials <- 1000L
n_experiments <- 2000L

theoretical_summary <- data.frame(
  Distribution = c(
    "Binomial, p = 0.5",
    "Beta-binomial, alpha = beta = 1",
    "Beta-binomial, alpha = beta = 20",
    "Beta-binomial, alpha = beta = 1e6"
  ),
  Mean = c(
    n_trials * 0.5,
    n_trials * 0.5,
    n_trials * 0.5,
    n_trials * 0.5
  ),
  Variance = c(
    n_trials * 0.5 * 0.5,
    n_trials * (2 * 1 + n_trials) / (4 * (2 * 1 + 1)),
    n_trials * (2 * 20 + n_trials) / (4 * (2 * 20 + 1)),
    n_trials * (2 * 1e6 + n_trials) / (4 * (2 * 1e6 + 1))
  )
)

knitr::kable(theoretical_summary, digits = 3)
```

```{r}
rbeta_binom <- function(n, size, alpha, beta) {
  p <- rbeta(n, shape1 = alpha, shape2 = beta)
  rbinom(n, size = size, prob = p)
}

samples <- list(
  "Binomial p = 0.5" =
    rbinom(n_experiments, size = n_trials, prob = 0.5),
  "Beta-binomial alpha = beta = 1" =
    rbeta_binom(n_experiments, size = n_trials, alpha = 1, beta = 1),
  "Beta-binomial alpha = beta = 20" =
    rbeta_binom(n_experiments, size = n_trials, alpha = 20, beta = 20),
  "Beta-binomial alpha = beta = 1e6" =
    rbeta_binom(n_experiments, size = n_trials, alpha = 1e6, beta = 1e6)
)

old_par <- par(mfrow = c(2, 2), mar = c(4.2, 4.2, 3, 1))
for (nm in names(samples)) {
  hist(
    samples[[nm]],
    breaks = 30,
    main = nm,
    xlab = "Number of successes",
    ylab = "Number of experiments",
    col = "#8FB9A8",
    border = "white",
    cex.main = 0.9
  )
}
par(old_par)
```

\newpage

## Problem 31: Likelihood of binary genotypes given a tumor mutation tree

### Part 1

For an attachment of cell $j$ to node $v$, the expected genotype has mutation $i$ present exactly when mutation $i$ lies on the path from the root to $v$. If $x_{ij}$ is the observed binary genotype, then

$$
P(x_{ij}\mid z_i=0)
= \alpha^{x_{ij}}(1-\alpha)^{1-x_{ij}},
$$

and

$$
P(x_{ij}\mid z_i=1)
= (1-\beta)^{x_{ij}}\beta^{1-x_{ij}},
$$

where $\alpha$ is the false positive rate and $\beta$ is the false negative rate.

For each cell, start with the log likelihood under the all absent genotype. When traversing the tree from parent to child, add only the log-likelihood increment for the mutation at the child node. Thus every mutation cell pair is processed a constant number of times, giving $O(IJ)$ time.

```{r}
##' computes the likelihood of the input genotype matrix given a tree
##' returns both the node-by-cell log likelihood, and the overall log likelihood
##' after marginalizing over all cell attachment
##' assume uniform attachment probabilities
##' @param tree list with $root (integer node index) and $children
##' (a list mapping each parent node index, as a string, to an integer vector
##' of its children's indices,
##' e.g. list(root=1L, children=list("1"=2L, "2"=c(3L,4L))))
##' @param genotype_matrix mutation-by-cell binary matrix of observed genotypes
##' @param alpha false positive rate
##' @param beta false negative rate
compute_log_lik <- function(tree, genotype_matrix, alpha, beta) {

  num_mutations <- nrow(genotype_matrix)
  num_cells <- ncol(genotype_matrix)

  log_lik <- matrix(
    NA_real_,
    nrow = num_mutations,
    ncol = num_cells,
    dimnames = list(
      paste0("node_", seq_len(num_mutations)),
      paste0("cell_", seq_len(num_cells))
    )
  )

  log_absent <- ifelse(genotype_matrix == 1, log(alpha), log1p(-alpha))
  log_present <- ifelse(genotype_matrix == 1, log1p(-beta), log(beta))

  absent_baseline <- colSums(log_absent)
  present_increment <- log_present - log_absent

  stack <- tree$root
  log_lik[tree$root, ] <- absent_baseline + present_increment[tree$root, ]

  while (length(stack) > 0) {
    parent <- stack[[length(stack)]]
    stack <- stack[-length(stack)]

    children <- tree$children[[as.character(parent)]]
    if (is.null(children)) {
      next
    }

    for (child in children) {
      log_lik[child, ] <- log_lik[parent, ] + present_increment[child, ]
      stack <- c(stack, child)
    }
  }

  if (anyNA(log_lik)) {
    stop("The tree does not reach every mutation node in the genotype matrix.")
  }

  cell_log_lik <- vapply(seq_len(num_cells), function(cell_idx) {
    node_log_lik <- log_lik[, cell_idx]
    max_log_lik <- max(node_log_lik)
    max_log_lik + log(mean(exp(node_log_lik - max_log_lik)))
  }, numeric(1))

  final_llik <- sum(cell_log_lik)

  return(list(log_lik = log_lik, final_llik = final_llik))
}
```

### Part 2

```{r}
genotype_matrix <- as.matrix(read.table("cells.txt"))
colnames(genotype_matrix) <- paste0("cell_", seq_len(ncol(genotype_matrix)))
rownames(genotype_matrix) <- paste0("mut_", seq_len(nrow(genotype_matrix)))
genotype_matrix
```

The provided file contains 5 mutations and 8 cells. The problem statement mentions 7 cells, but the matrix in cells.txt has 8 columns. We use the data as given. This matrix does not permit a perfect phylogeny without errors. In a perfect phylogeny under the infinite sites assumption, for any pair of mutations, the sets of cells carrying those mutations must be nested or disjoint. Here, for example,

$$
\begin{aligned}
S_2 &= \{1,2,4,5,6,7,8\},\\
S_3 &= \{1,2,3\}.
\end{aligned}
$$

These two sets overlap in cells $\{1,2\}$, but neither set contains the other. Similarly, $S_4=\{6,7,8\}$ and $S_5=\{5,6\}$ overlap without being nested. Such crossing mutation patterns cannot be represented by a perfect phylogeny with each mutation occurring exactly once and never being lost.

### Parts 3 and 4

Use the following mutation trees.

```{r}
tree1 <- list(
  root     = 1L,
  children = list(
    "1" = 2L,
    "2" = c(3L, 4L),
    "4" = 5L
  )
)

tree2 <- list(
  root     = 1L,
  children = list(
    "1" = 2L,
    "2" = c(3L, 4L, 5L)
  )
)
```

```{r}
cases <- list(
  "T1: alpha=0.1, beta=0.01" =
    list(tree = tree1, alpha = 0.1, beta = 0.01),
  "T2: alpha=0.1, beta=0.01" =
    list(tree = tree2, alpha = 0.1, beta = 0.01),
  "T1: alpha=0.01, beta=0.1" =
    list(tree = tree1, alpha = 0.01, beta = 0.1),
  "T2: alpha=0.01, beta=0.1" =
    list(tree = tree2, alpha = 0.01, beta = 0.1)
)

results <- lapply(cases, function(case) {
  compute_log_lik(
    tree = case$tree,
    genotype_matrix = genotype_matrix,
    alpha = case$alpha,
    beta = case$beta
  )
})

best_attachment <- lapply(results, function(result) {
  apply(result$log_lik, 2, which.max)
})

result_table <- do.call(rbind, lapply(names(results), function(nm) {
  data.frame(
    Case = nm,
    Log_likelihood = results[[nm]]$final_llik,
    Best_node_by_cell = paste(best_attachment[[nm]], collapse = ", ")
  )
}))
row.names(result_table) <- NULL

knitr::kable(result_table, digits = 3)
```

The heatmaps below show the node-by-cell log likelihoods returned by compute_log_lik. Larger values indicate more likely posthoc attachment of the corresponding cell to that node.

```{r}
#| fig-width: 8
#| fig-height: 8
#| out-width: "100%"

plot_log_lik_heatmap <- function(log_lik, main) {
  num_nodes <- nrow(log_lik)
  num_cells <- ncol(log_lik)
  z <- t(log_lik[num_nodes:1, , drop = FALSE])

  image(
    x = seq_len(num_cells),
    y = seq_len(num_nodes),
    z = z,
    col = hcl.colors(80, "Viridis"),
    axes = FALSE,
    xlab = "Cells (observed genotypes)",
    ylab = "Node (expected tree profiles)",
    main = main
  )
  axis(1, at = seq_len(num_cells), labels = seq_len(num_cells))
  axis(2, at = seq_len(num_nodes), labels = num_nodes:1, las = 1)
  box()

  text(
    x = rep(seq_len(num_cells), times = num_nodes),
    y = rep(seq_len(num_nodes), each = num_cells),
    labels = sprintf("%.1f", as.vector(z)),
    cex = 0.55,
    col = "white"
  )
}

old_par <- par(mfrow = c(2, 2), mar = c(4.2, 4.2, 3, 1))
for (nm in names(results)) {
  plot_log_lik_heatmap(results[[nm]]$log_lik, nm)
}
par(old_par)
```


When false positives are more common than false negatives (when $\alpha > \beta$), $(\alpha,\beta)=(0.1,0.01)$, Tree 2 has the higher marginal log likelihood:

$$
-17.957 > -18.490.
$$

This setting makes it cheap to explain extra observed 1's as false positives. Tree 2 therefore fits the data better because mutation 5 is a sibling of mutations 3 and 4, allowing cell 5 to attach directly to node 5 without forcing mutation 4 to be present.

When false negatives are more common than false positives (when $\beta > \alpha$), $(\alpha,\beta)=(0.01,0.1)$, Tree 1 has the higher marginal log likelihood:

$$
-19.362 > -21.106.
$$

This setting makes missing observed 1's less costly, but extra observed 1's now cost $\log\alpha=\log(0.01)\approx -4.6$ each. Tree 1 has the ancestry $4\rightarrow 5$, so cell 6, which carries both mutations 4 and 5, fits node 5 of Tree 1 perfectly. In Tree 2, mutations 4 and 5 are siblings, so no single node carries both, and cell 6 must incur at least one false positive under either attachment. This penalty is what tips the balance toward Tree 1 when false positives are expensive. Thus the preferred tree and the maximum-likelihood cell attachments change with the error parameters.
