---
title: "Homework 2: EM Algorithm and Motif Finding"
author: "Bernhard Buss, Timothy Lauren Nijhuis, Carlo Rossetto and Görkem Kadir Solun"
format: pdf
device: "cairo_pdf"

editor: 
  markdown: 
    wrap: 72
editor_options: 
  chunk_output_type: inline
---

```{r}
set.seed(42)
library(patchwork)
library(ggplot2)
```

## (4) TSG inactivation problem formulation

### (4a) The complete-data log-likelihood

$$\log(P(X,Z|\lambda,\theta)) = \sum_{i = 1}^{N}\log(P(X^{(i)},Z^{(i)}|\lambda,\theta) = \sum_{i = 1}^{N}\sum_{k = 1}^{2}Z^{(ik)}\log(\lambda_{k}P(X^{(i)}|\theta_{k}))$$
$P(X^{(i)}|\theta_k)$ follows a Bernoulli distribution with probability
$\theta$ for each position upto position L. So
$$ P(X^{(i)}|\theta_{k}) = \prod_{j = 1}^{L}\theta_{k}^{X^{(i)}}(1-\theta_{k})^{(1 - X^{(i)})}$$
Substituting
$$\log(P(X,Z|\lambda,\theta)) = \sum_{i = 1}^{N}\sum_{k = 1}^{2}Z^{(ik)}\log(\lambda_{k}\prod_{j = 1}^{L}\theta_{k}^{X^{(i)}}(1-\theta_{k})^{(1 - X^{(i)})}) $$
$$= \sum_{i = 1}^{N}\sum_{k = 1}^{2}Z^{(ik)}\left(\log(\lambda_{k})+\prod_{j = 1}^{L}\left(X^{(i)}\log(\theta_{k})+(1 - X^{(i)})\log(1-\theta_{k})\right)\right) $$
$$ = \sum_{i = 1}^{N}Z^{(i1)}\left(\log(\lambda_{1})+\prod_{j = 1}^{L}\left(X^{(i)}\log(\theta_{1})+(1 - X^{(i)})\log(1-\theta_{1})\right)\right) + Z^{(i2)}\left(\log(\lambda_{2})+\prod_{j = 1}^{L}\left(X^{(i)}\log(\theta_{2})+(1 - X^{(i)})\log(1-\theta_{2})\right)\right) $$

### (4b) The responsibility

$$\gamma_{ik} = P(Z^{(ik)}=1|X,\lambda,\theta)$$ Using Bayes theorem:
$$ P(G|C) = \frac{P(G)P(C|G)}{P(C)} $$ with \$ C = Z\^{(ik)} = 1\$ and
\$ G = X\$. Resulting in

$$ P(Z^{(ik)}=1|X,\lambda,\theta) = \frac{P(Z^{(ik)}=1|\lambda,\theta)P(X^{(i)}|\lambda,\theta)}{P(X^{(i)}|\lambda,\theta)} $$
$$ = \frac{\lambda_{k}\prod_{j = 1}^{L}\theta_{k}^{X^{(i)}}(1-\theta_{k})^{(1 - X^{(i)})}}{P(X^{(i)}|\lambda,\theta)} $$
\$ P(X\^{(i)}\|\lambda,\theta) \$ is the marginal probability of
$X^{(i)}$, so one needs to sum over $Z^{(ik)} = 1$.
$$ P(X^{(i)}|\lambda,\theta) = \sum_{j = 1}^{2}\lambda_{k}\prod_{j = 1}^{L}\theta_{k}^{X^{(i)}}(1-\theta_{k})^{(1 - X^{(i)})}$$
Resulting in:
$$ \gamma_{ik} =\frac{\lambda_{k}\prod_{j = 1}^{L}\theta_{k}^{X^{(i)}}(1-\theta_{k})^{(1 - X^{(i)})}}{\sum_{k = 1}^{2}\lambda_{k}\prod_{j = 1}^{L}\theta_{k}^{X^{(i)}}(1-\theta_{k})^{(1 - X^{(i)})}} $$

### (4c) The M-step update formula

## (5) TSG inactivation problem implementation

### (5a) The EM algorithm

Please write your code in the indicated blocks.

```{r}
##' This function implements the EM algorithm for the gene mutation problem
##' @param D Data matrix of dimensions N-by-100, where N is the number of observations
##' @param k Number of cell groups
##' @param lambda Vector of mixture weights
##' @param theta Vector of probabilities of mutations
##' @param tolerance A threshold used to check convergence
mutation_EM <- function(D, k, lambda, theta, tolerance = 1e-2) {
  
  # observed log-likelihood
  ll_obs <- -Inf
  # difference between two iterations
  diff <- Inf
  # number of observations
  N <- nrow(D)
  # responsibilities
  gamma <- matrix(0, nrow = k, ncol = N)
  # keep track of lambda and theta during the optimization
  lambda_all <- lambda
  theta_all <- theta
  # iteration number
  t <- 1


  # run the E-step and M-step until convergence 
  while (diff > tolerance) {
    
    ############# E-step #############
    
    ### YOUR CODE STARTS ###
    
    # Compute the responsibilities
    # ...

    # Update observed log-likelihood 
    # ll_obs_new <- ...
    ll_obs <- c(ll_obs, ll_obs_new) 
    # keep track of ll_obs throughout iterations

    # Recompute difference between two iterations
    diff <- abs(ll_obs[t] - ll_obs[t+1])

    
    ############# M-step #############
    
    # Recompute priors (mixture weights)
    # ... 

    lambda_all <- rbind(lambda_all, lambda)
    # keep track of lambda throughout iterations
    
    # Recompute probability of mutation for each cell group
    # ...
    
    theta_all <- rbind(theta_all, theta)
    # keep track of theta throughout iterations

    t <- t+1
  }
  ### YOUR CODE ENDS ###
  
  return(list(ll_obs = ll_obs, lambda = lambda, theta = theta, gamma = gamma, 
              lambda_all = lambda_all, theta_all = theta_all))
  
}
```

### (2b) Load data, initialize parameters, and run the EM algorithm.

```{r}
# read the data into D
D <- read.csv("hw2_student/mutations.csv")
D <- as.matrix(D)
# check the dimension of D
all(dim(D) == c(200, 100))

# number of cell groups
k <- 2
# mixture weights (a vector of length k)
lambda <- c(0.5, 0.5)
# probabilities of mutation (having a 1 in the data) (a vector of length k)
theta <- c(0.6, 0.4)

# run the EM algorithm
res <- mutation_EM(D, k, lambda, theta)
```

### (2c) Plotting convergence.

Observed log-likelihood throughout iterations:

```{r}
## YOUR CODE ##
```

Probabilities of mutation throughout iterations:

```{r}
## YOUR CODE ##
```

Mixture weights throughout iterations:

```{r}
## YOUR CODE ##
```

### (2d) Responsibilities at final iteration:

Heatmap of responsibilities:

```{r}
## YOUR CODE ##
```

How many cells belong to the mutated group, and how many cells belong to
the wild-type group?
