---
title: "Homework 2: EM Algorithm and Motif Finding"
author: "Bernhard Buss, Timothy Lauren Nijhuis, Carlo Rossetto and Görkem Kadir Solun"
format: pdf
device: "cairo_pdf"

editor: 
  markdown: 
    wrap: 72
editor_options: 
  chunk_output_type: inline
---

```{r}
set.seed(42)
library(patchwork)
library(ggplot2)
```

## (4) TSG inactivation problem formulation

### (4a) The complete-data log-likelihood

$$\log(P(X,Z|\lambda,\theta)) = \sum_{i = 1}^{N}\log(P(X^{(i)},Z^{(i)}|\lambda,\theta) = \sum_{i = 1}^{N}\sum_{k = 1}^{2}Z^{(ik)}\log(\lambda_{k}P(X^{(i)}|\theta_{k}))$$
$P(X^{(i)}|\theta_k)$ follows a Bernoulli distribution with probability
$\theta$ for each position upto position $L$. So
$$ P(X^{(i)}|\theta_{k}) = \prod_{l = 1}^{L}\theta_{k}^{X_{l}^{(i)}}(1-\theta_{k})^{\left(1 - X_{l}^{(i)}\right)}$$
Substituting
$$\log(P(X,Z|\lambda,\theta)) = \sum_{i = 1}^{N}\sum_{k = 1}^{2}Z^{(ik)}\log\left(\lambda_{k}\prod_{l = 1}^{L}\theta_{k}^{X_{l}^{(i)}}(1-\theta_{k})^{\left(1 - X_{l}^{(i)}\right)}\right) $$
$$= \sum_{i = 1}^{N}\sum_{k = 1}^{2}Z^{(ik)}\left(\log(\lambda_{k})+\prod_{l = 1}^{L}\left(X_{l}^{(i)}\log(\theta_{k})+(1 - X_{l}^{(i)})\log(1-\theta_{k})\right)\right) $$
$$ = \sum_{i = 1}^{N}Z^{(i1)}\left(\log(\lambda_{1})+\prod_{l = 1}^{L}\left(X_{l}^{(i)}\log(\theta_{1})+(1 - X_{l}^{(i)})\log(1-\theta_{1})\right)\right) + Z^{(i2)}\left(\log(\lambda_{2})+\prod_{l = 1}^{L}\left(X_{l}^{(i)}\log(\theta_{2})+(1 - X_{l}^{(i)})\log(1-\theta_{2})\right)\right) $$

### (4b) The responsibility

$$\gamma_{ik} = P(Z^{(ik)}=1|X,\lambda,\theta)$$ Using Bayes theorem:
$$ P(G|C) = \frac{P(G)P(C|G)}{P(C)} $$ with $C = Z^{(ik)} = 1$ and
$G = X$. Resulting in

$$ P(Z^{(ik)}=1|X,\lambda,\theta) = \frac{P(Z^{(ik)}=1|\lambda,\theta)P(X^{(i)}|\lambda,\theta)}{P(X^{(i)}|\lambda,\theta)} $$
$$ = \frac{\lambda_{k}\prod_{l = 1}^{L}\theta_{k}^{X_{l}^{(i)}}(1-\theta_{k})^{\left(1 - X_{l}^{(i)}\right)}}{P(X^{(i)}|\lambda,\theta)} $$
$P(X^{(i)}|\lambda,\theta)$ is the marginal probability of $X^{(i)}$, so
one needs to sum over $Z^{(ik)} = 1$.
$$ P(X^{(i)}|\lambda,\theta) = \sum_{j = 1}^{2}\lambda_{k}\prod_{l = 1}^{L}\theta_{k}^{X_{l}^{(i)}}(1-\theta_{k})^{\left(1 - X_{l}^{(i)}\right)}$$
Resulting in:
$$ \gamma_{ik} =\frac{\lambda_{k}\prod_{l = 1}^{L}\theta_{k}^{X_{l}^{(i)}}(1-\theta_{k})^{\left(1 - X_{l}^{(i)}\right)}}{\sum_{k = 1}^{2}\lambda_{k}\prod_{l = 1}^{L}\theta_{k}^{X_{l}^{(i)}}(1-\theta_{k})^{\left(1 - X_{l}^{(i)}\right)}}  $$

### (4c) The M-step update formula

The M-step maximise $F(q,\theta)$ with respect to $\theta$:

$$
F(q,\theta) =E_{q}[l_{hid}(\theta)] - E_{q}[\log(q(Z))]
$$

Since only $E_{q}[l_{hid}(\theta)]$ depends on $\theta$,
$E_{q}[l_{hid}(\theta)]$ is maximised. For a motif mixture model, as
shown in the lectures, the complete data likelihood can be further
simplified to:
$$ E_{q}[l_{hid}(\theta)] = \sum_{i,k}\gamma_{ik}\log(\lambda_k) + \sum_{i,k}\gamma_{ik}\log(P(X^{(i)}|\theta_k)) $$

Maximisation with respect to the $\lambda_k$ involves only the left sum
term.
$$\mu \left( 1 - \sum_{k=1}^2 \lambda_k \right) + \sum_{i=1}^N \sum_{k=1}^2 \gamma_{ik} \log \lambda_k$$
where $\mu$ is a Lagrange multiplier for the constraint
$\sum_{k=1}^2 \lambda_k = 1$. The update formula for $\lambda_k$ is
obtained by setting the derivative of this objective to zero:
$$\frac{\sum_i \gamma_{ik}}{\lambda_k} - \mu = 0 \quad \text{for each } k, \quad \text{and } \sum_{k=1}^2 \lambda_k = 1.$$
Rearranging, we have that
$$\sum_i \gamma_{ik} = \mu \cdot \lambda_k, \quad \text{for each } k.$$
Summing the two equations from $k$ equals 1 and 2 together, and using
the fact that by definition $\sum_{k=1}^2 \gamma_{ik} = 1$ for each $i$
and $\sum_{k=1}^2 \lambda_k = 1$, we have that $\mu = N$, and
$$\lambda_{k}^{new} = \frac{\sum_i \gamma_{ik}}{N} = \frac{\sum_i \frac{\lambda_{k}\prod_{l = 1}^{L}\theta_{k}^{X_{l}^{(i)}}(1-\theta_{k})^{\left(1 - X_{l}^{(i)}\right)}}{\sum_{k = 1}^{2}\lambda_{k}\prod_{l = 1}^{L}\theta_{k}^{X_{l}^{(i)}}(1-\theta_{k})^{\left(1 - X_{l}^{(i)}\right)}}}{N}$$
where $N$ is the total number of cells.

Regarding maximising for $\theta_k$, it is the right sum that should be
maximised (because it is the sum that depends on $\theta_k$).

Since from (4a)
$$ P(X^{(i)}|\theta_{k}) = \prod_{l = 1}^{L}\theta_{k}^{X_{l}^{(i)}}(1-\theta_{k})^{\left(1 - X_{l}^{(i)}\right)} $$

The right sum becomes

$$\sum_{i,k}\gamma_{ik}\log(P(X^{(i)}|\theta_k)) = \sum_{i,k}\gamma_{ik}\log\left(\prod_{l = 1}^{L}\theta_{k}^{X_{l}^{(i)}}(1-\theta_{k})^{\left(1 - X_{l}^{(i)}\right)}\right) =$$
$$ \sum_{i,k}\gamma_{ik}\sum_{l = 1}^{L}\left(X_{l}^{(i)}\log(\theta_{k})+(1 - X_{l}^{(i)})\log(1-\theta_{k})\right)$$

For ease the expression above is set $=f$. In order to find the optimal
$\theta$ the first derivative is set $=0$.

$$ f = \sum_{i,k}\gamma_{ik}\sum_{l = 1}^{L}\left(X_{l}^{(i)}\log(\theta_{k})+(1 - X_{l}^{(i)})\log(1-\theta_{k})\right) =\sum_{i}^{N}\sum_{k=1}^{2}\gamma_{ik}\sum_{l = 1}^{L}\left(X_{l}^{(i)}\log(\theta_{k})+(1 - X_{l}^{(i)})\log(1-\theta_{k})\right) $$

When taking the derivative, the $\sum_{k = 1}^{2}$ falls out

$$
\frac{\partial{f}}{\partial{\theta_k}} = 0 = \sum_{i}\gamma_{ik}\sum_{l = 1}^{L}\left(\frac{X_{l}^{(i)}}{\theta_{k}}-\frac{1 - X_{l}^{(i)}}{1-\theta_{k}}\right) $$\
$$ = \sum_{i}\gamma_{ik}\sum_{l = 1}^{L}\left(\frac{X_{l}^{(i)}(1-\theta_{k})-\theta_{k}(1 - X_{l}^{(i)})}{\theta_{k}(1-\theta_{k})}\right) = \sum_{i}\gamma_{ik}\sum_{l = 1}^{L}\left(\frac{X_{l}^{(i)}-\theta_{k}}{\theta_{k}(1-\theta_{k})}\right)
$$

$$
\implies \sum_{i}\gamma_{ik}\sum_{l = 1}^{L}\left(\frac{\theta_{k}}{\theta_{k}(1-\theta_{k})}\right) = \sum_{i}\gamma_{ik}\sum_{l = 1}^{L}\left(\frac{X_{l}^{(i)}}{\theta_{k}(1-\theta_{k})}\right)
$$

$$
\implies \left(\frac{\theta_{k}}{\theta_{k}(1-\theta_{k})}\right) \sum_{i}\gamma_{ik}\sum_{l = 1}^{L}1 = \left(\frac{1}{\theta_{k}(1-\theta_{k})}\right)\sum_{i}\gamma_{ik}\sum_{l = 1}^{L}X_{l}^{(i)}
$$

$$
\implies \theta_{k}\cdot L\sum_{i}\gamma_{ik} = \sum_{i}\gamma_{ik}\sum_{l = 1}^{L}X_{l}^{(i)} $$ 
$$\implies \theta_{k}^{new} = \frac{\sum_{i}\gamma_{ik}\sum_{l = 1}^{L}X_{l}^{(i)}}{L\sum_{i}\gamma_{ik}}
$$

## (5) TSG inactivation problem implementation

### (5a) The EM algorithm

Please write your code in the indicated blocks.

```{r}
##' This function implements the EM algorithm for the gene mutation problem
##' @param D Data matrix of dimensions N-by-100, where N is the number of observations
##' @param k Number of cell groups
##' @param lambda Vector of mixture weights
##' @param theta Vector of probabilities of mutations
##' @param tolerance A threshold used to check convergence
mutation_EM <- function(D, k, lambda, theta, tolerance = 1e-2) {
  
  # observed log-likelihood
  ll_obs <- -Inf
  # difference between two iterations
  diff <- Inf
  # number of observations
  N <- nrow(D)
  # responsibilities
  gamma <- matrix(0, nrow = k, ncol = N)
  # keep track of lambda and theta during the optimization
  lambda_all <- lambda
  theta_all <- theta
  # iteration number
  t <- 1


  # run the E-step and M-step until convergence 
  while (diff > tolerance) {
    
    ############# E-step #############
    
    ### YOUR CODE STARTS ###
    
    # Compute the responsibilities
    # ...

    # Update observed log-likelihood 
    # ll_obs_new <- ...
    ll_obs <- c(ll_obs, ll_obs_new) 
    # keep track of ll_obs throughout iterations

    # Recompute difference between two iterations
    diff <- abs(ll_obs[t] - ll_obs[t+1])

    
    ############# M-step #############
    
    # Recompute priors (mixture weights)
    # ... 

    lambda_all <- rbind(lambda_all, lambda)
    # keep track of lambda throughout iterations
    
    # Recompute probability of mutation for each cell group
    # ...
    
    theta_all <- rbind(theta_all, theta)
    # keep track of theta throughout iterations

    t <- t+1
  }
  ### YOUR CODE ENDS ###
  
  return(list(ll_obs = ll_obs, lambda = lambda, theta = theta, gamma = gamma, 
              lambda_all = lambda_all, theta_all = theta_all))
  
}
```

### (2b) Load data, initialize parameters, and run the EM algorithm.

```{r}
# read the data into D
D <- read.csv("hw2_student/mutations.csv")
D <- as.matrix(D)
# check the dimension of D
all(dim(D) == c(200, 100))

# number of cell groups
k <- 2
# mixture weights (a vector of length k)
lambda <- c(0.5, 0.5)
# probabilities of mutation (having a 1 in the data) (a vector of length k)
theta <- c(0.6, 0.4)

# run the EM algorithm
res <- mutation_EM(D, k, lambda, theta)
```

### (2c) Plotting convergence.

Observed log-likelihood throughout iterations:

```{r}
## YOUR CODE ##
```

Probabilities of mutation throughout iterations:

```{r}
## YOUR CODE ##
```

Mixture weights throughout iterations:

```{r}
## YOUR CODE ##
```

### (2d) Responsibilities at final iteration:

Heatmap of responsibilities:

```{r}
## YOUR CODE ##
```

How many cells belong to the mutated group, and how many cells belong to
the wild-type group?
