---
title: "Project 2 EM Algorithm"
output: pdf
---

```{r}
set.seed(42)
library(patchwork)
library(ggplot2)
```

## (a) The EM algorithm

Please write your code in the indicated blocks.


```{r}
##' This function implements the EM algorithm for the gene mutation problem
##' @param D Data matrix of dimensions N-by-100, where N is the number of observations
##' @param k Number of cell groups
##' @param lambda Vector of mixture weights
##' @param theta Vector of probabilities of mutations
##' @param tolerance A threshold used to check convergence
mutation_EM <- function(D, k, lambda, theta, tolerance = 1e-2) {
  
  # observed log-likelihood
  ll_obs <- -Inf
  # difference between two iterations
  diff <- Inf
  # number of observations
  N <- nrow(D)
  # responsibilities
  gamma <- matrix(0, nrow = k, ncol = N)
  # keep track of lambda and theta during the optimization
  lambda_all <- lambda
  theta_all <- theta
  # iteration number
  t <- 1


  # run the E-step and M-step until convergence 
  while (diff > tolerance) {
    
    ############# E-step #############
    
    ### YOUR CODE STARTS ###
    
    # Compute the responsibilities
    # ...

    # Update observed log-likelihood 
    # ll_obs_new <- ...
    ll_obs <- c(ll_obs, ll_obs_new) 
    # keep track of ll_obs throughout iterations

    # Recompute difference between two iterations
    diff <- abs(ll_obs[t] - ll_obs[t+1])

    
    ############# M-step #############
    
    # Recompute priors (mixture weights)
    # ... 

    lambda_all <- rbind(lambda_all, lambda)
    # keep track of lambda throughout iterations
    
    # Recompute probability of mutation for each cell group
    # ...
    
    theta_all <- rbind(theta_all, theta)
    # keep track of theta throughout iterations

    t <- t+1
  }
  ### YOUR CODE ENDS ###
  
  return(list(ll_obs = ll_obs, lambda = lambda, theta = theta, gamma = gamma, 
              lambda_all = lambda_all, theta_all = theta_all))
  
}
```


## (b) Load data, initialize parameters, and run the EM algorithm.

```{r}
# read the data into D
D <- read.csv("hw2_student/mutations.csv")
D <- as.matrix(D)
# check the dimension of D
all(dim(D) == c(200, 100))

# number of cell groups
k <- 2
# mixture weights (a vector of length k)
lambda <- c(0.5, 0.5)
# probabilities of mutation (having a 1 in the data) (a vector of length k)
theta <- c(0.6, 0.4)

# run the EM algorithm
res <- mutation_EM(D, k, lambda, theta)
```


## (c) Plotting convergence.

Observed log-likelihood throughout iterations:

```{r}
## YOUR CODE ##
```

Probabilities of mutation throughout iterations:

```{r}
## YOUR CODE ##
```

Mixture weights throughout iterations:

```{r}
## YOUR CODE ##
```


## (D) Responsibilities at final iteration:

Heatmap of responsibilities:

```{r}
## YOUR CODE ##
```

How many cells belong to the mutated group, and how many cells belong to the wild-type group?
