---
title: "Advanced R tutorial: EM algorithm"
date: "February 25, 2026"
format: pdf
---

## Motif finding

We assume that reads consisting of several of the bases A,C,G,T (1,2,3,4) come from either a background distribution (independent of position) or an alternative.

First, we have to define the ground truth parameters. They contain the probabilities of the alternative and the mixture weight.

```{r}
## load plotting libraries
library(ggplot2)
library(patchwork)

## random parameters
set.seed(2024)

# generate probabilities of sampling nucleotides
phi1 <- matrix(runif(10*4, 0, 1), 4)
phi1 <- apply(phi1, 2, function(x) x/sum(x))
phi2 <- runif(4, 0, 1)
phi2 <- phi2/sum(phi2)

# generate mixing weights
mw <- runif(1, 0.1, 0.9)

```

In the next part, we generate data (reads of length n=10) based on our ground truth parameters. 

```{r}
## generate data
nreads <- 100
data1 <- numeric()
for (i in 1:10) {  
  # iterate over alignment position
    tmp <- numeric()
    for (j in 1:round(nreads*mw)) {  
      # iterate over reads of the motif component 
        tmp <- c(tmp, sample(1:4, 1, prob = phi1[, i])) 
        # sample base
    }
    data1 <- cbind(data1, tmp) 
}
data0 <- numeric()
for (i in 1:10) { 
  # iterate over alignment position
    tmp <- numeric()
    for (j in 1:round(nreads*(1-mw))) {  
      # iterate over reads of the background component 
        tmp <- c(tmp, sample(1:4, 1, prob = phi2))
    }
    data0 <- cbind(data0, tmp)
}
data <- rbind(data1, data0)
```

## plot sequences

```{r}
data_long <- reshape2::melt(unname(data), varnames = c("Seq", "Position"), 
                            value.name = "base")

# Plot
ggplot(data_long, aes(x = Position, y = Seq, fill = factor(base))) +
  geom_tile() +
  geom_hline(yintercept = round(nreads*mw), linetype = "dashed") +  
  # Add horizontal dashed line at y = mw
  labs(x = "Position", y = "Sequence Index", fill = "Base") +
  theme_minimal()
```

We pretend to neither know the model parameters nor which component generated which read. We use an expectation maximization algorithm to iteratively compute the expectation of the reads and the model parameters until convergence.

```{r}
## implementation of the EM
em <- function(data) {
    ## random inits
    m1 <- matrix(runif(4*ncol(data), 0, 1), 4)  
    # parameters for nucleotide frequencies in the motif
    m1 <- apply(m1, 2, function(x) x/sum(x))
    m0 <- runif(4, 0, 1) 
    # parameters for nucleotide frequencies in the background
    m0 <- m0/sum(m0)
    lambda <- runif(1) 
    # mixture weights
    ## initial E-step
    probs <- matrix(0, nrow(data), 2) 
    # store responsibilities 
    for (i in 1:nrow(data)) { 
      # iterate over sequences
        tmp <- tmp0 <- 1
        for (j in 1:ncol(data)) { 
          # iterate over alignment columns
            ## product over the probabilities of each data point of a sample 
            #given each component
            tmp <- m1[data[i, j], j]*tmp   
            # access the rows of m1 which is equal to data[i, j], and its jth column 
            tmp0 <- m0[data[i, j]]*tmp0  
            # access the row of m0 which is equal to data[i, j]
        }
        probs[i, 1] <- tmp*lambda 
        # multiply by mixture weights to obtain unnormalised responsibilities
        probs[i, 2] <- tmp0*(1-lambda)
    }
    ## calculate the observed data log likelihood of the model
    lls <- ll <- sum(log(apply(probs, 1, sum)))
    probs <- t(apply(probs, 1, function(x) x/sum(x))) 
    # normalise the responsibilities 
    probs_all <- probs[,1] 
    # to keep track of this quantity 
    lambda <- apply(probs, 2, sum) 
    # compute mixture weights
    lambdas <- lambda <- lambda[1]/sum(lambda)
    llold <- log(0)
    count <- 0
    while(count < 100) {  
      # iterate the EM
        count <- count + 1
        llold <- ll
        lambda_old <- lambda
        ## M-step
        m0 <- 0 
        for (i in 1:ncol(data)) { 
          # iterate over positions see slide 20
            ## calculate new parameters for nucleotide frequencies in the motif 
            # with weighted probabilities
            m1[, i] <- c(sum(probs[which(data[, i] == 1), 1]), 
                         # <- sum the probs of the first component 
                         # (column index == 1), 
                         sum(probs[which(data[, i] == 2), 1]), 
                         # where row index corresponds to the nucleotide 
                         sum(probs[which(data[, i] == 3), 1]), 
                         # appearing in the data at that position 
                         sum(probs[which(data[, i] == 4), 1])) 
            m0 <- m0 + c(sum(probs[which(data[, i] == 1), 2]),
                          sum(probs[which(data[, i] == 2), 2]),
                          sum(probs[which(data[, i] == 3), 2]),
                          sum(probs[which(data[, i] == 4), 2]))
        }
        m1 <- apply(m1, 2, function(x) x/sum(x)) 
        # normalise
        m0 <- m0/sum(m0)
        ## E-step, as above
        probs <- matrix(0, nrow(data), 2)
        for (i in 1:nrow(data)) { 
          # iterate over sequences
            tmp <- tmp0 <- 1
            for (j in 1:ncol(data)) { 
              # iterate over alignment columns
                ## product over the probabilities of each data point of a sample 
                # given each component
                tmp <- m1[data[i, j], j]*tmp  
                # access the rows of m1 which is equal to data[i, j], and its jth column 
                tmp0 <- m0[data[i, j]]*tmp0 
                # access the row of m0 which is equal to data[i, j]
            }
            probs[i, 1] <- tmp*lambda 
            # multiply by mixture weights to obtain unnormalised responsibilities
            probs[i, 2] <- tmp0*(1-lambda)
        }
        ## calculate the observed data log likelihood of the model
        ll <- sum(log(apply(probs, 1, sum)))
        lls <- c(lls, ll)
        probs <- t(apply(probs, 1, function(x) x/sum(x))) 
        # normalise the responsibilities 
        lambda <- apply(probs, 2, sum)  
        # compute mixture weights, see slide 19
        lambda <- lambda[1]/sum(lambda)
        lambdas <- c(lambdas, lambda)
        probs_all <- cbind(probs_all, probs[,1])
    }
    return(list(phi1 = m1, phi2 = m0, lambda = lambda, ll = ll, lls = lls, 
                lambdas = lambdas, probs = probs, probs_all = probs_all))
}
```

We put the data into our our function and let it run. To assess the result, we plot the ground truth against our inferred values.

```{r}
res <- em(data)

## plot correlation between parameters

# Create a data frame for ggplot
plot_data <- data.frame(
  x = c(as.vector(phi1), phi2),
  y = c(as.vector(res$phi1), res$phi2)
)

# Calculate correlation
cor_value <- cor(plot_data$x, plot_data$y)

# Plot using ggplot
ggplot(plot_data, aes(x = x, y = y)) +
  geom_point() + # Plot points
  geom_abline(intercept = 0, slope = 1, color = "red") + 
  # Add y=x line
  ggtitle(paste("Correlation:", round(cor_value, 3))) + 
  # Add correlation as title
  theme_light() + 
  # Use a light theme
  xlab("True parameters") + 
  # Label for x-axis
  ylab("Inferred parameters") 
# Label for y-axis


## convergence rate of LLS and lambda
plot_data <- data.frame(Step = 1:length(res$lls), LLS = res$lls, lambdas = res$lambdas)

# plot lls
plt1 <- ggplot(plot_data, aes(x = Step, y = LLS)) +
  geom_line() + 
  labs(
    title = "Convergence of the log likelihood",
    x = "Step",
    y = "LLS Value"
  ) +
  theme_light() 


## plot  of lambda
plt2 <- ggplot(plot_data, aes(x = Step, y = lambdas)) +
  geom_line() + 
  labs(
    title = "Convergence of lambda",
    x = "Step",
    y = "LLS Value"
  ) +
  geom_hline(yintercept = mw, linetype = "dashed", color = "blue") +  
  # Add horizontal dashed line at y = mw
  annotate("text", x = Inf, y = mw, label = "Groundtruth", vjust = 0,
           hjust = 1, color = "blue") +  
  # Add label
  theme_light() +
  ylim(c(0,1))

plt1 + plt2

## plot progression of sequence assignment
data_long <- reshape2::melt(t(unname(res$probs_all)), 
                            varnames = c("Time", "Sequence"), 
                            value.name = "Probability")

# Plot
ggplot(data_long, aes(x = Time, y = Sequence, fill = Probability)) +
  geom_tile() +
  scale_fill_gradient(low = "blue", high = "red") +
  geom_hline(yintercept = round(nreads*mw), linetype = "dashed") +  
  # Add horizontal dashed line at y = mw
  labs(x = "Time", y = "Sequence Index", fill = "P(Component 1)") +
  theme_minimal()



```
