---
title: "Homework 3: Markov chains and hidden Markov models"
author:
  - Bernhard Buss
  - Dario Franzen
  - Aurimas Greicius
  - Timothy Lauren Nijhuis
  - Carlo Rossetto
  - Yuliyan Slavchev
  - Görkem Kadir Solun
format: pdf
device: "cairo_pdf"

editor: 
  markdown: 
    wrap: 72
editor_options: 
  chunk_output_type: inline
---

```{r}
set.seed(42)
```

## (6) Identify the most likely coding versus non-coding composition of a given DNA sequence

### (6a) The Viterbi quantities for sequence TAAC

All probabilities mentioned are $\log_{10}$-transformed.

The initial state probabilities (I), transition probabilities (T) and
emission probabilities (E), that will be used to calculate the Viterbi
quantities, $v_{k}(n)$, are shown below.

| Initial state probabilities (I) | C-Region | NC-Region |
|---------------------------------|----------|-----------|
|                                 | -0.4     | -0.2      |

| Transition probabilities (T) | C-Region | NC-Region |
|------------------------------|----------|-----------|
| C-Region                     | -0.50    | -0.15     |
| NC-Region                    | -0.20    | -0.40     |

| Emission probabilities (E) | A    | C    | G    | T    |
|----------------------------|------|------|------|------|
| C-Region                   | -0.7 | -0.5 | -1.0 | -0.4 |
| NC-Region                  | -0.7 | -0.2 | -1.0 | -1.0 |

For T at position 1, we should compute $P_{C-Region}(T,1)$ and
$P_{NC-Region}(T,1)$ or in another notation $v_{C-Region}(1)$ and
$v_{NC-Region}(1)$ for T observed.

$$ P_{C-Region}(T,1) = I_{C-Region} + E_{C-Region,T} = -0.4 - 0.4 = -0.8 $$
$$ P_{NC-Region}(T,1) = I_{NC-Region} + E_{NC-Region,T} = -0.2 - 1.0 = -1.2 $$

For A at position 2, we should compute $P_{C-Region}(A,2)$ and
$P_{NC-Region}(A,2)$ or in another notation $v_{C-Region}(2)$ and
$v_{NC-Region}(2)$ for A observed.

$$ P_{C-Region}(A,2) = E_{C-Region,A} + \max
\begin{cases} P_{C-Region}(T,1) + T_{C-Region,C-Region} \\ P_{NC-Region}(T,1)+T_{NC-Region,C-Region} \end{cases} $$

$$ =-0.7 + \max \begin{cases} -0.8  -0.5 \\ -1.2 - 0.2 \end{cases} = -0.7 + \max \begin{cases} -1.3 \\ -1.4 \end{cases} = -2.0 $$

$$ P_{NC-Region}(A,2) = E_{NC-Region,A} + \max \begin{cases} P_{NC-Region}(T,1) + T_{NC-Region,NC-Region}\\ P_{C-Region}(T,1)+T_{C-Region,NC-Region} \end{cases} $$

$$ =-0.7 + \max \begin{cases} -1.2  -0.4 \\ -0.8 - 0.15 \end{cases} = -0.7 + \max \begin{cases} -1.60 \\ -0.95 \end{cases} = -1.65 $$

$P_{C-Region}(A,2)$ comes from position 1 C-region. $P_{NC-Region}(A,2)$
comes from position 1 C-region.

For A at position 3, we should compute $P_{C-Region}(A,3)$ and
$P_{NC-Region}(A,3)$ or in another notation $v_{C-Region}(3)$ and
$v_{NC-Region}(3)$ for A observed.

$$ P_{C-Region}(A,3) = E_{C-Region,A} + \max
\begin{cases} P_{C-Region}(A,2) + T_{C-Region,C-Region}\\ P_{NC-Region}(A,2)+T_{NC-Region,C-Region} \end{cases} $$

$$ =-0.7 + \max \begin{cases} -2.0  -0.5 \\ -1.65 - 0.2 \end{cases} = -0.7 + \max \begin{cases} -2.50 \\ -1.85 \end{cases} = -2.55 $$

$$ P_{NC-Region}(A,3) = E_{NC-Region,A} + \max \begin{cases} P_{NC-Region}(A,2) + T_{NC-Region,NC-Region}\\ P_{C-Region}(A,2)+T_{C-Region,NC-Region} \end{cases} $$

$$ =-0.7 + \max \begin{cases} -1.65  -0.4 \\ -2.0 - 0.15 \end{cases} = -0.7 + \max \begin{cases} -2.05 \\ -2.15 \end{cases} = -2.75 $$

$P_{C-Region}(A,3)$ comes from position 2 NC-region.
$P_{NC-Region}(A,3)$ comes from position 2 NC-region.

For C at position 4, we should compute $P_{C-Region}(C,4)$ and
$P_{NC-Region}(C,4)$ or in another notation $v_{C-Region}(4)$ and
$v_{NC-Region}(4)$ for C observed.

$$ P_{C-Region}(C,4) = E_{C-Region,C} + \max
\begin{cases} P_{C-Region}(A,3) + T_{C-Region,C-Region}\\ P_{NC-Region}(A,3)+T_{NC-Region,C-Region} \end{cases} $$

$$ =-0.5 + \max \begin{cases} -2.55  -0.5 \\ -2.75 - 0.2 \end{cases} = -0.5 + \max \begin{cases} -3.05 \\ -2.95 \end{cases} = -3.45 $$

$$ P_{NC-Region}(C,4) = E_{NC-Region,A} + \max \begin{cases} P_{NC-Region}(A,3) + T_{NC-Region,NC-Region}\\ P_{C-Region}(A,3)+T_{C-Region,NC-Region} \end{cases} $$

$$ =-0.2 + \max \begin{cases} -2.75  -0.4 \\ -2.55 - 0.15 \end{cases} = -0.2 + \max \begin{cases} -3.15 \\ -2.70 \end{cases} = -2.9 $$

$P_{C-Region}(C,4)$ comes from position 3 NC-region.
$P_{NC-Region}(C,4)$ comes from position 3 C-region.

The resulting Viterbi quantities, ($log_{10}$-transformed probabilities)
are summarised in the table below.

|           | T     | A     | A     | C     |
|-----------|-------|-------|-------|-------|
| C-Region  | -0.80 | -2.00 | -2.55 | -3.45 |
| NC-Region | -1.20 | -1.65 | -2.75 | -2.90 |

The pointers (previous state visited) are summarised in the table below.

| Pointers | T | A | A | C |
|---------------|---------------|---------------|---------------|---------------|
| C-Region | Not applicable | Position 1, C region | Position 2, NC region | Position 3, NC region |
| NC-Region | Not applicable | Position 1, C region | Position 2, NC region | Position 3, C region |

### (6b) The most likely hidden path

The state in the last position, where C is observed, with the highest
probability is NC-Region, with a probability of -2.9. Using the pointers
summarised in the previous tables, the resulting shortest path is: C,
NC, C, NC, or more precisely summarised in table format below:

| Shortest path         | Position 1 | Position 2 | Position 3 | Position 4 |
|-----------------------|------------|------------|------------|------------|
| Hidden variable state | C-Region   | NC-Region  | C-Region   | NC-Region  |

## (7) Predicting protein secondary structure using HMMs

### (7a) Reading and storing data

Saving the tab-delimited files as dataframes

```{r}
proteins_train <- read.delim("exercise/data/proteins_train.tsv",sep="\t", header = FALSE)
proteins_test <- read.delim("exercise/data/proteins_test.tsv",sep="\t", header = FALSE)
proteins_new <- read.delim("exercise/data/proteins_new.tsv",sep="\t", header = FALSE)
```

Renaming the columns.

```{r}
colnames(proteins_train) <- c("Protein ID","Amino acid sequence","Secondary structure code")
colnames(proteins_test) <- c("Protein ID","Amino acid sequence","Secondary structure code")
colnames(proteins_new) <- c("Protein ID","Amino acid sequence")
```

### (7b) Initial state probabilities (I), transition probabilities (T) and emission probababilities (E)

```{r}
L <- nrow(proteins_train)

aa_codes <- c("A","C","D","E","F","G","H","I","K","L","M","N","P","Q","R","S","T","U","V","W","X","Y")
  
ss_codes <- c("B","C","E","G","H","I","S","T")

I_counts <- matrix(0, nrow = length(ss_codes), ncol = 1, dimnames = list(ss_codes))
T_counts <- matrix(0, nrow = length(ss_codes),ncol = length(ss_codes),dimnames = list(ss_codes,ss_codes))
E_counts <- matrix(0, nrow = length(ss_codes),ncol = length(aa_codes),dimnames = list(ss_codes,aa_codes))

for (protein in 1:L) {
 
  aa_seq <- proteins_train$`Amino acid sequence`[protein]
  ss_seq <- proteins_train$`Secondary structure code`[protein]
  
  # initialization counts, C is the secondary structure of the first amino acid
  first_ss <- substr(ss_seq,1,1)
  I_counts[first_ss,] <- I_counts[first_ss,] + 1
  
  l_seq = nchar(aa_seq)
  
  for (N in 1:l_seq) {
    aa_N <- substr(aa_seq,N,N)
    ss_N <- substr(ss_seq,N,N)
    
    # emission counts
    E_counts[ss_N,aa_N] <- E_counts[ss_N,aa_N] + 1
    
    #transmission counts
    if (N != 1) {
      T_counts[prev_ss_N,ss_N] <- T_counts[prev_ss_N,ss_N] + 1
    }
    prev_ss_N <- ss_N
  }
}

# maximum likelihood estimates
I <- I_counts/sum(I_counts)
E <- E_counts/rowSums(E_counts)
T <- T_counts/rowSums(T_counts)

print(I)
print(T)
print(E)
```

### (7c) The stationary distribution $\pi$

An ergodic Markov chain has a unique stationary distribution
$\pi = (\pi_l)_{l \in [K]}$ such that

$$\lim_{n \to \infty} T_{kl}^n = \pi_l = \sum_{k \in [K]} \pi_k T_{kl}, \quad l \in [K], \quad \sum_{l \in [K]} \pi_l = 1$$

independent of the initial distribution $I$. In our situation:
$$ \sum_{k = 1}^{8} \pi_{k} = 1 $$

In matrix notation, $\pi$ is the solution of $\pi^t = \pi^t T$. $\pi$ is
the eigenvector of the transpose of T corresponding to an eigenvalue of
1.

```{r}
decomposed_T <- eigen(t(T))
idx_value1 <- which.min(abs(decomposed_T$values - 1))
vector_value1 <- decomposed_T$vectors[,idx_value1]

pi <- vector_value1/sum(vector_value1)
pi <- Re(pi)
pi <- pi / sum(pi)
names(pi) <- ss_codes

pi_matrix = matrix(pi, dimnames = list(ss_codes))
print((pi_matrix))

pi_next <- as.numeric(pi %*% T)
names(pi_next) <- ss_codes

print(all.equal(pi, pi_next))
print(cbind(Original_pi = pi, After_One_Step = pi_next))
```

### (7d) The latent state sequence Z using Viterbi algorithm

```{r}
source("exercise/code/viterbi.r")

log_I <- log(I)
log_T <- log(T)
log_E <- log(E)

# for proteins_test
proteins_test_pred <- proteins_test
colnames(proteins_test_pred) <- c("ProteinID","AminoAcids","SecondaryStructure")

proteins_test_result <- viterbi(
  E = log_E,
  Tr = log_T,
  I = log_I,
  p = proteins_test_pred
)

# proteins_test_result as tsv file
write.table(proteins_test_result, file="proteins_test_pred.tsv", 
            quote=FALSE, sep="\t", col.names = FALSE, row.names = FALSE)

# for proteins_new
proteins_new_pred <- proteins_new
colnames(proteins_new_pred) <- c("ProteinID","AminoAcids")

proteins_new_result <- viterbi(
  E = log_E,
  Tr = log_T,
  I = log_I,
  p = proteins_new_pred
)

# protein_new_result as tsv file
write.table(proteins_new_result, file="proteins_new_pred.tsv", 
            quote=FALSE, sep="\t", col.names = FALSE, row.names = FALSE)
```

### (7e) Confidence intervals via bootstrapping

```{r}
B <- 1000
n_proteins <- nrow(proteins_train)

estimate_params <- function(df, ss_codes, aa_codes) {
  n_rows <- nrow(df)
  I_counts <- matrix(0, nrow = length(ss_codes), ncol = 1, dimnames = list(ss_codes, "Init"))
  T_counts <- matrix(0, nrow = length(ss_codes), ncol = length(ss_codes), dimnames = list(ss_codes, ss_codes))
  E_counts <- matrix(0, nrow = length(ss_codes), ncol = length(aa_codes), dimnames = list(ss_codes, aa_codes))

  for (protein in 1:n_rows) {
    aa_seq <- df$`Amino acid sequence`[protein]
    ss_seq <- df$`Secondary structure code`[protein]

    first_ss <- substr(ss_seq, 1, 1)
    I_counts[first_ss, ] <- I_counts[first_ss, ] + 1

    l_seq <- nchar(aa_seq)
    if (l_seq > 0) {
      aa_split <- strsplit(aa_seq, "")[[1]]
      ss_split <- strsplit(ss_seq, "")[[1]]

      for (N in 1:l_seq) {
        E_counts[ss_split[N], aa_split[N]] <- E_counts[ss_split[N], aa_split[N]] + 1
        if (N > 1) {
          T_counts[ss_split[N - 1], ss_split[N]] <- T_counts[ss_split[N - 1], ss_split[N]] + 1
        }
      }
    }
  }

  I_probs <- I_counts / sum(I_counts)
  row_sums_E <- rowSums(E_counts)
  E_probs <- ifelse(row_sums_E == 0, 0, E_counts / row_sums_E)
  row_sums_T <- rowSums(T_counts)
  T_probs <- ifelse(row_sums_T == 0, 0, T_counts / row_sums_T)

  list(I = I_probs, T = T_probs, E = E_probs)
}

I_boot <- array(0, dim = c(length(ss_codes), 1, B), dimnames = list(ss_codes, "Init", NULL))
T_boot <- array(0, dim = c(length(ss_codes), length(ss_codes), B), dimnames = list(ss_codes, ss_codes, NULL))
E_boot <- array(0, dim = c(length(ss_codes), length(aa_codes), B), dimnames = list(ss_codes, aa_codes, NULL))

set.seed(42)

for (b in 1:B) {
  # Sample with replacement
  resampled_indices <- sample(1:n_proteins, n_proteins, replace = TRUE)
  boot_data <- proteins_train[resampled_indices, ]

  params <- estimate_params(boot_data, ss_codes, aa_codes)
  I_boot[, , b] <- params$I
  T_boot[, , b] <- params$T
  E_boot[, , b] <- params$E
}

# Compute the empirical 95% confidence intervals (2.5% and 97.5% quantiles)
# We use the apply() function over the 1st and 2nd dimensions (rows and columns)
I_lower <- apply(I_boot, c(1, 2), quantile, probs = 0.025, na.rm = TRUE)
I_upper <- apply(I_boot, c(1, 2), quantile, probs = 0.975, na.rm = TRUE)

T_lower <- apply(T_boot, c(1, 2), quantile, probs = 0.025, na.rm = TRUE)
T_upper <- apply(T_boot, c(1, 2), quantile, probs = 0.975, na.rm = TRUE)

E_lower <- apply(E_boot, c(1, 2), quantile, probs = 0.025, na.rm = TRUE)
E_upper <- apply(E_boot, c(1, 2), quantile, probs = 0.975, na.rm = TRUE)

dimnames(I_lower) <- list(ss_codes, "Init")
dimnames(I_upper) <- list(ss_codes, "Init")
dimnames(T_lower) <- list(ss_codes, ss_codes)
dimnames(T_upper) <- list(ss_codes, ss_codes)
dimnames(E_lower) <- list(ss_codes, aa_codes)
dimnames(E_upper) <- list(ss_codes, aa_codes)

# To verify
print(I_lower)
print(I_upper)
print(T_lower)
print(T_upper)
print(E_lower)
print(E_upper)
```

### (7f) Accuracy of predictions

```{r}
compute_accuracy <- function(pred_seq, true_seq) {
  pred_chars <- strsplit(pred_seq, "")[[1]]
  true_chars <- strsplit(true_seq, "")[[1]]
  l_seq <- length(true_chars)

  if (l_seq == 0) {
    return(NA_real_)
  }

  sum(pred_chars == true_chars) / l_seq
}

n_test_proteins <- nrow(proteins_test_result)
accuracies <- numeric(n_test_proteins)

for (i in 1:n_test_proteins) {
  real_seq <- proteins_test_result[i, 3]
  pred_seq <- proteins_test_result[i, 4]
  accuracies[i] <- compute_accuracy(pred_seq, real_seq)
}

summary(accuracies)
```

The accuracies vary significantly across the protein sequences, from
around 0% (min) to 86 % (maximum). The median and mean of 32% indicate
there is room for improvement.
