---
title: "Homework 3: Markov chains and hidden Markov models"
author: "Bernhard Buss, Timothy Lauren Nijhuis, Carlo Rossetto and Görkem Kadir Solun"
format: pdf
device: "cairo_pdf"

editor: 
  markdown: 
    wrap: 72
editor_options: 
  chunk_output_type: inline
---

```{r}
set.seed(42)
library(patchwork)
library(ggplot2)
```

## (6) Identify the most likely coding versus non-coding composition of a given DNA sequence

### (6a) The Viterbi quantities for sequence TAAC

All probabilities mentioned are $\log_{10}$-transformed.

The initial state probabilities (I), transition probabilities (T) and
emission probabilities (E), that will be used to calculate the Viterbi
quantities, $v_{k}(n)$, are shown below.

| Initial state probabilities (I) | C-Region | NC-Region |
|---------------------------------|----------|-----------|
|                                 | -0.4     | -0.2      |

| Transition probabilities (T) | C-Region | NC-Region |
|------------------------------|----------|-----------|
| C-Region                     | -0.50    | -0.15     |
| NC-Region                    | -0.20    | -0.40     |

| Emission probabilities (E) | A    | C    | G    | T    |
|----------------------------|------|------|------|------|
| C-Region                   | -0.7 | -0.5 | -1.0 | -0.4 |
| NC-Region                  | -0.7 | -0.2 | -1.0 | -1.0 |

For T at position 1, we should compute $P_{C-Region}(T,1)$ and
$P_{NC-Region}(T,1)$ or in another notation $v_{C-Region}(1)$ and
$v_{NC-Region}(1)$ for T observed.

$$ P_{C-Region}(T,1) = I_{C-Region} + E_{C-Region,T} = -0.4 - 0.4 = -0.8 $$
$$ P_{NC-Region}(T,1) = I_{NC-Region} + E_{NC-Region,T} = -0.2 - 1.0 = -1.2 $$

For A at position 2, we should compute $P_{C-Region}(A,2)$ and
$P_{NC-Region}(A,2)$ or in another notation $v_{C-Region}(2)$ and
$v_{NC-Region}(2)$ for A observed.

$$ P_{C-Region}(A,2) = E_{C-Region,A} + \max
\begin{cases} P_{C-Region}(T,1) + T_{C-Region,C-Region} \\ P_{NC-Region}(T,1)+T_{NC-Region,C-Region} \end{cases} $$

$$ =-0.7 + \max \begin{cases} -0.8  -0.5 \\ -1.2 - 0.2 \end{cases} = -0.7 + \max \begin{cases} -1.3 \\ -1.4 \end{cases} = -2.0 $$

$$ P_{NC-Region}(A,2) = E_{NC-Region,A} + \max \begin{cases} P_{NC-Region}(T,1) + T_{NC-Region,NC-Region}\\ P_{C-Region}(T,1)+T_{C-Region,NC-Region} \end{cases} $$

$$ =-0.7 + \max \begin{cases} -1.2  -0.4 \\ -0.8 - 0.15 \end{cases} = -0.7 + \max \begin{cases} -1.60 \\ -0.95 \end{cases} = -1.65 $$

$P_{C-Region}(A,2)$ comes from position 1 C-region. $P_{NC-Region}(A,2)$
comes from position 1 C-region.

For A at position 3, we should compute $P_{C-Region}(A,3)$ and
$P_{NC-Region}(A,3)$ or in another notation $v_{C-Region}(3)$ and
$v_{NC-Region}(3)$ for A observed.

$$ P_{C-Region}(A,3) = E_{C-Region,A} + \max
\begin{cases} P_{C-Region}(A,2) + T_{C-Region,C-Region}\\ P_{NC-Region}(A,2)+T_{NC-Region,C-Region} \end{cases} $$

$$ =-0.7 + \max \begin{cases} -2.0  -0.5 \\ -1.65 - 0.2 \end{cases} = -0.7 + \max \begin{cases} -2.50 \\ -1.85 \end{cases} = -2.55 $$

$$ P_{NC-Region}(A,3) = E_{NC-Region,A} + \max \begin{cases} P_{NC-Region}(A,2) + T_{NC-Region,NC-Region}\\ P_{C-Region}(A,2)+T_{C-Region,NC-Region} \end{cases} $$

$$ =-0.7 + \max \begin{cases} -1.65  -0.4 \\ -2.0 - 0.15 \end{cases} = -0.7 + \max \begin{cases} -2.05 \\ -2.15 \end{cases} = -2.75 $$

$P_{C-Region}(A,3)$ comes from position 2 NC-region.
$P_{NC-Region}(A,3)$ comes from position 2 NC-region.

For C at position 4, we should compute $P_{C-Region}(C,4)$ and
$P_{NC-Region}(C,4)$ or in another notation $v_{C-Region}(4)$ and
$v_{NC-Region}(4)$ for C observed.

$$ P_{C-Region}(C,4) = E_{C-Region,C} + \max
\begin{cases} P_{C-Region}(A,3) + T_{C-Region,C-Region}\\ P_{NC-Region}(A,3)+T_{NC-Region,C-Region} \end{cases} $$

$$ =-0.5 + \max \begin{cases} -2.55  -0.5 \\ -2.75 - 0.2 \end{cases} = -0.5 + \max \begin{cases} -3.05 \\ -2.95 \end{cases} = -3.45 $$

$$ P_{NC-Region}(C,4) = E_{NC-Region,A} + \max \begin{cases} P_{NC-Region}(A,3) + T_{NC-Region,NC-Region}\\ P_{C-Region}(A,3)+T_{C-Region,NC-Region} \end{cases} $$

$$ =-0.2 + \max \begin{cases} -2.75  -0.4 \\ -2.55 - 0.15 \end{cases} = -0.2 + \max \begin{cases} -3.25 \\ -2.70 \end{cases} = -2.9 $$

$P_{C-Region}(C,4)$ comes from position 3 NC-region.
$P_{NC-Region}(C,4)$ comes from position 3 C-region.

The resulting Viterbi quantities, ($log_{10}$-transformed probabilities) are
summarised in the table below.

|           | T     | A     | A     | C     |
|-----------|-------|-------|-------|-------|
| C-Region  | -0.80 | -2.00 | -2.55 | -3.45 |
| NC-Region | -1.20 | -1.65 | -2.75 | -2.90 |

The pointers (previous state visited) are summarised in the table below.

| Pointers | T | A | A | C |
|---------------|---------------|---------------|---------------|---------------|
| C-Region | Not applicable | Position 1, C region | Position 2, NC region | Position 3, NC region |
| NC-Region | Not applicable | Position 1, C region | Position 2, NC region | Position 3, C region |

### (6b) The most likely hidden path

The state in the last position, where C is observed, with the highest probability is NC-Region, with a probability of -2.9. Using the pointers summarised in the previous tables, the resulting shortest path is: C, NC, C, NC, or more precisely summarised in table format below:

|Shortest path           | Position 1| Position 2|Position 3|Position 4|
|-----------|-------|-------|-------|-------|
| Hidden variable state  | C-Region | NC-Region | C-Region | NC-Region |

## (7) Predicting protein secondary structure using HMMs

### (7a) Reading and storing data

Complete the implementation of the EM algorithm in the code skeleton.
Note: Beware of numerical instability when multiplying large amount of
numbers together – do additions of logarithms whenever possible.

Please write your code in the indicated blocks.

```{r}
##' This function implements the EM algorithm for the gene mutation problem
##' @param D Data matrix of dimensions N-by-100, where N is the number of observations
##' @param k Number of cell groups
##' @param lambda Vector of mixture weights
##' @param theta Vector of probabilities of mutations
##' @param tolerance A threshold used to check convergence
mutation_EM <- function(D, k, lambda, theta, tolerance = 1e-2) {
  
  # observed log-likelihood
  ll_obs <- -Inf
  # difference between two iterations
  diff <- Inf
  # number of observations
  N <- nrow(D)
  # number of positions
  L <- ncol(D)
  # responsibilities
  gamma <- matrix(0, nrow = k, ncol = N)
  # keep track of lambda and theta during the optimization
  lambda_all <- lambda
  theta_all <- theta
  # iteration number
  t <- 1

  # run the E-step and M-step until convergence 
  while (diff > tolerance) {
    
    ############# E-step #############
    
    ### YOUR CODE STARTS ###
    
    # Compute the responsibilities using log-space to avoid underflow
    # For each observation i and component j, compute
    # log(lambda_j) + sum_l [ X_l^(i) * log(theta_j) + (1 - X_l^(i)) * log(1 - theta_j) ]
    log_rho <- matrix(0, nrow = k, ncol = N)
    for (j in 1:k) {
      # Sum of X_l * log(theta_j) + (1-X_l) * log(1-theta_j) for each observation
      log_lik <- D %*% rep(log(theta[j]), L) + (1 - D) %*% rep(log(1 - theta[j]), L)
      # log(lambda_j) + log(P(X^(i) | theta_j))
      log_rho[j, ] <- log(lambda[j]) + as.vector(log_lik)
    }
    
    # Normalise responsibilities using log-sum-exp trick
    log_sum <- apply(log_rho, 2, function(x) {
      m <- max(x)
      m + log(sum(exp(x - m)))
    })
    for (j in 1:k) {
      gamma[j, ] <- exp(log_rho[j, ] - log_sum)
    }

    # Update observed log-likelihood
    # ll_obs_new = sum_i log( sum_j lambda_j * P(X^(i) | theta_j) )
    ll_obs_new <- sum(log_sum)
    ll_obs <- c(ll_obs, ll_obs_new) 
    # keep track of ll_obs throughout iterations

    # Recompute difference between two iterations
    diff <- abs(ll_obs[t] - ll_obs[t+1])

    
    ############# M-step #############
    
    # Recompute priors (mixture weights)
    # lambda_j^new = sum_i gamma_ij / N
    for (j in 1:k) {
      lambda[j] <- sum(gamma[j, ]) / N
    }

    lambda_all <- rbind(lambda_all, lambda)
    # keep track of lambda throughout iterations
    
    # Recompute probability of mutation for each cell group
    # theta_j^new = sum_i gamma_ij * sum_l X_l^(i) / (L * sum_i gamma_ij)
    for (j in 1:k) {
      numerator <- sum(gamma[j, ] * rowSums(D))
      denominator <- L * sum(gamma[j, ])
      theta[j] <- numerator / denominator
    }
    
    theta_all <- rbind(theta_all, theta)
    # keep track of theta throughout iterations

    t <- t+1
  }
  ### YOUR CODE ENDS ###
  
  return(list(ll_obs = ll_obs, lambda = lambda, theta = theta, gamma = gamma, 
              lambda_all = lambda_all, theta_all = theta_all))
  
}
```

### (7b) The vector of initial state probabilities I

Use the given code to read in the data stored in the file
‘mutations.csv’, and to initialize the mixture weights λ and the
mutation probabilities θ. We assume a priori that the mixture weights
are uniform, and θ_1 \> θ_2. Run the completed EM algorithm on this
data.

```{r}
# read the data into D
D <- read.csv("mutations.csv")
D <- as.matrix(D)
# check the dimension of D
all(dim(D) == c(200, 100))

# number of cell groups
k <- 2
# mixture weights (a vector of length k)
lambda <- c(0.5, 0.5)
# probabilities of mutation (having a 1 in the data) (a vector of length k)
theta <- c(0.6, 0.4)

# run the EM algorithm
res <- mutation_EM(D, k, lambda, theta)
```

### (7c) The stationary distribution $\pi$

Plot the following quantities (y-axis) as a function of the number of
iterations (x-axis) in three separate plots: • the observed
log-likelihood (recall that this is computed by marginalizing out the
hidden variables) • the mutation probabilities θ_1 and θ_2, and • the
mixture weights λ_1 and λ_2.

Observed log-likelihood throughout iterations:

```{r}
# Remove the initial -Inf for plotting
ll_data <- data.frame(
  Iteration = 1:(length(res$ll_obs) - 1),
  LogLikelihood = res$ll_obs[-1]
)

ggplot(ll_data, aes(x = Iteration, y = LogLikelihood)) +
  geom_line() +
  geom_point(size = 1.5) +
  labs(title = "Observed Log-Likelihood Convergence",
       x = "Iteration", y = "Observed Log-Likelihood") +
  theme_light()
```

Probabilities of mutation throughout iterations:

```{r}
theta_data <- data.frame(
  Iteration = 0:(nrow(res$theta_all) - 1),
  Theta1 = res$theta_all[, 1],
  Theta2 = res$theta_all[, 2]
)
theta_long <- rbind(
  data.frame(Iteration = theta_data$Iteration, Component = "Theta1", Theta = theta_data$Theta1),
  data.frame(Iteration = theta_data$Iteration, Component = "Theta2", Theta = theta_data$Theta2)
)

ggplot(theta_long, aes(x = Iteration, y = Theta, color = Component)) +
  geom_line() +
  geom_point(size = 1.5) +
  labs(title = "Mutation Probabilities Throughout Iterations",
       x = "Iteration", y = expression(theta)) +
  scale_color_manual(values = c("Theta1" = "red", "Theta2" = "blue"),
                     labels = c(expression(theta[1]~"(mutated)"),
                                expression(theta[2]~"(wild-type)"))) +
  theme_light()
```

Mixture weights throughout iterations:

```{r}
lambda_data <- data.frame(
  Iteration = 0:(nrow(res$lambda_all) - 1),
  Lambda1 = res$lambda_all[, 1],
  Lambda2 = res$lambda_all[, 2]
)
lambda_long <- rbind(
  data.frame(Iteration = lambda_data$Iteration, Component = "Lambda1", Lambda = lambda_data$Lambda1),
  data.frame(Iteration = lambda_data$Iteration, Component = "Lambda2", Lambda = lambda_data$Lambda2)
)

ggplot(lambda_long, aes(x = Iteration, y = Lambda, color = Component)) +
  geom_line() +
  geom_point(size = 1.5) +
  labs(title = "Mixture Weights Throughout Iterations",
       x = "Iteration", y = expression(lambda)) +
  scale_color_manual(values = c("Lambda1" = "red", "Lambda2" = "blue"),
                     labels = c(expression(lambda[1]~"(mutated)"),
                                expression(lambda[2]~"(wild-type)"))) +
  ylim(0, 1) +
  theme_light()
```

### (7d) The latent state sequence Z

Plot a heatmap of the responsibilities γ at the final step, where
columns are sequences, rows are components, and values indicate the
responsibility of component k for X\^(i). For each cell, we assign it to
the component with higher probability. How many cells belong to the
mutated group, and how many cells belong to the wild-type group?

Heatmap of responsibilities:

```{r}
# gamma is k x N; rows = components, columns = sequences
# Sort cells by responsibility to the mutated component for clearer visualization
cell_order <- order(res$gamma[1, ], decreasing = TRUE)
gamma_sorted <- res$gamma[, cell_order]

resp_data <- rbind(
  data.frame(Component = "Mutated", Cell = 1:ncol(gamma_sorted),
             Responsibility = gamma_sorted[1, ]),
  data.frame(Component = "Wild-type", Cell = 1:ncol(gamma_sorted),
             Responsibility = gamma_sorted[2, ])
)

ggplot(resp_data, aes(x = Cell, y = Component, fill = Responsibility)) +
  geom_tile() +
  scale_fill_gradient(low = "blue", high = "red",
                      name = expression(gamma[ik])) +
  labs(title = "Responsibilities at Final Iteration (sorted by mutated responsibility)",
       x = "Cell Index (sorted)", y = "Component") +
  theme_minimal()
```

```{r}
# gamma is k x N; rows = components, columns = sequences
resp_data <- rbind(
  data.frame(Component = "Mutated", Cell = 1:ncol(gamma_sorted),
             Responsibility = res$gamma[1, ]),
  data.frame(Component = "Wild-type", Cell = 1:ncol(gamma_sorted),
             Responsibility = res$gamma[2, ])
)

ggplot(resp_data, aes(x = Cell, y = Component, fill = Responsibility)) +
  geom_tile() +
  scale_fill_gradient(low = "blue", high = "red",
                      name = expression(gamma[ik])) +
  labs(title = "Responsibilities at Final Iteration (not sorted by mutated responsibility)",
       x = "Cell Index", y = "Component") +
  theme_minimal()
```

How many cells belong to the mutated group, and how many cells belong to
the wild-type group?

```{r}
# Assign each cell to the component with the highest responsibility
n_mutated <- sum(res$gamma[1, ] > 0.5)
n_wildtype <- sum(res$gamma[1, ] <= 0.5)
cat("Number of cells in the mutated group:", n_mutated, "\n")
cat("Number of cells in the wild-type group:", n_wildtype, "\n")
```

### (7e) Confidence intervals

### (7f) Accuracy of predictions
