---
title: "Hw3"
format: html
editor: visual
---

## Problem 7

\(a\) Read proteins train.tsv, proteins test.tsv and proteins new.tsv into the memory and store each in a data.frame.

```{r}
library(readr)
library(dplyr)

proteins_train <- read_tsv("~/Desktop/StatsModelsCompBio/Week3/SMCB Markov Models Homework/exercise/data/proteins_train.tsv", col_names=FALSE)
proteins_test <- read_tsv("~/Desktop/StatsModelsCompBio/Week3/SMCB Markov Models Homework/exercise/data/proteins_test.tsv", col_names=FALSE)
proteins_new <- read_tsv("~/Desktop/StatsModelsCompBio/Week3/SMCB Markov Models Homework/exercise/data/proteins_new.tsv", col_names=FALSE)

states_vocab <- c("B", "C", "E", "G", "H", "I", "S", "T")
amino_vocab <- c("A", "C", "D", "E", "F", "G", "H", "I", "K", "L", "M", "N", "P", "Q", "R", "S", "T", "U", "V", "W", "X", "Y")
```

\(b\) Estimate the vector of initial state probabilities I, the matrix of transition probabilities T and the matrix for emission probabilities E by using maximum likelihood, i.e., by counting how often each transition and each emission happened and how often each secondary structure was found at the beginning of a protein, respectively.

```{r}

# For estimating initial probabilities, we can simply count of starting letter / total seqs to get the probabilistic estimate:

# amino_vocab
# states_vocab
# proteins_train$X3

# Create a dictionary
states_dict <- setNames(rep(0, length(states_vocab)), states_vocab)

# Loop over each sequence in the dataset
for (seq in proteins_train$X3) {
  
  # Extract just the very first character of the sequence
  first_char <- substr(seq, 1, 1)
  
  # Add 1 to the corresponding dictionary entry
  if (first_char %in% names(states_dict)) {
    states_dict[first_char] <- states_dict[first_char] + 1
  }
}

# Divide all entries by total sum to get probabilities
initial_probs <- states_dict / sum(states_dict)

print("Initial Probabilities (I):")
print(initial_probs)

# For emissions, we calculate the X3 -> X2 counts, divide by total count for that specific state to get prob.

E_counts <- matrix(0, nrow = length(states_vocab), ncol = length(amino_vocab))
rownames(E_counts) <- states_vocab
colnames(E_counts) <- amino_vocab

# For transitions, for each sequence we check X3_n -> X3_n+1 counts, divide by total count for that specific state

T_counts <- matrix(0, nrow = length(states_vocab), ncol = length(states_vocab))
rownames(T_counts) <- states_vocab # Current state
colnames(T_counts) <- states_vocab # Next state

# Loop over every sequence in your dataset
for (i in 1:nrow(proteins_train)) {
  
  # Split the sequences into individual characters
  state_seq <- strsplit(proteins_train$X3[i], "")[[1]]
  obs_seq   <- strsplit(proteins_train$X2[i], "")[[1]] 
  
  # --- EMISSIONS COUNTING ---
  for (n in 1:length(state_seq)) {
    s <- state_seq[n] # Current state
    o <- obs_seq[n]   # Current observation (amino acid)
    E_counts[s, o] <- E_counts[s, o] + 1
  }
  
  # --- TRANSITIONS COUNTING ---
  if (length(state_seq) > 1) {
    for (n in 1:(length(state_seq) - 1)) {
      s_curr <- state_seq[n]     # Current state
      s_next <- state_seq[n + 1] # Next state
      T_counts[s_curr, s_next] <- T_counts[s_curr, s_next] + 1
    }
  }
}

# rowSums() gives the total number of times each state occurred
E_probs <- E_counts / rowSums(E_counts)
T_probs <- T_counts / rowSums(T_counts)

```

\(c\) Estimate the stationary distribution π of the Markov chain. 

```{r}
# Eigen decomposition of the transpose of the transition matrix
eig <- eigen(t(T_probs))

# Find the index of the eigenvalue closest to 1
idx <- which.min(abs(eig$values - 1))

# Extract the corresponding eigenvector and normalize it so it sums to 1
pi_stat <- Re(eig$vectors[, idx])
pi_stat <- pi_stat / sum(pi_stat)
names(pi_stat) <- states_vocab

print("Stationary Distribution (pi):")
print(pi_stat)

# ---------------------------------------------------------
# Sanity check: Does multiplying by T change the distribution?
# ---------------------------------------------------------

# Multiply the stationary distribution (row vector) by your transition matrix
pi_next <- as.numeric(pi_stat %*% T_probs)
names(pi_next) <- states_vocab

print(all.equal(pi_stat, pi_next))

# Combine them into a table to see them side-by-side
comparison_table <- cbind(Original_pi = pi_stat, 
                          After_One_Step = pi_next)

print(comparison_table)
```

\(d\) Having estimated the parameters, i.e., the emission and transition matrices E,T and the vector of initial state probabilities I, you can predict the latent state sequence Z of a protein’s amino acid sequence X using the Viterbi algorithm. Use the Viterbi algorithm provided in viterbi.r (carefully read the parameter description!) and iterate over each data.frame of proteins test.tsv and proteins new.tsv row by row and use the amino acid sequence to predict its secondary structure, which you add to the data.frame as a new column. Save the extended data.frame of proteins new.tsv including the predicted secondary structure as a tsv file and hand it in together with your pdf.

```{r}
# source("viterbi.r")
#' @param E n times m matrix with the emission log probabilities of the n latent variables and m observed variables
#' @param Tr n times n matrix with the transition log probabilities
#' @param I numeric vector of length n with the initial log probabilities of the n latent variables
#' @param p a data.frame with a Variable named AminoAcids which holds the amino acid sequence as a character string
viterbi <- function(E, Tr, I, p) {
    .as.array <- function(.) stringr::str_split(., "")[[1]]
    unique.ss <- c("B", "C", "E", "G", "H", "I", "S", "T")
    unique.aa <- c("A", "C", "D", "E", "F", "G", "H", "I",
                   "K", "L", "M", "N", "P", "Q", "R", "S",
                   "T", "U", "V", "W", "X", "Y")
    for (k in seq(nrow(p))) {
        sequence <- p$AminoAcids[k]
        aa.vec <- .as.array(sequence) %>% match(unique.aa)
        P      <- matrix(0, nrow(E), length(aa.vec))
        Ptr    <- matrix(0, nrow(E), length(aa.vec))
        
        ## sets the paths
        for (i in seq(length(aa.vec))) {
            if (i == 1) {
                P[, i] <- I + E[, aa.vec[i]]
            } else {
                for (j in seq(nrow(E))) {
                    p.loc    <- P[, i - 1] + Tr[, j] + E[j, aa.vec[i]]
                    P[j, i] <- max(p.loc)
                    Ptr[j, i] <- which.max(p.loc)
                }
            }
        }
        
        ## backtrace: computes the most likely path
        Phi <- vector(mode="integer",   length=length(aa.vec))
        Phi[length(Phi)] <- which.max(P[, ncol(P)])
        ## we start at the back, just as with Needleman-Wunsch or Smith-Waterman
        for (i in seq(from=length(aa.vec), to=2)) {
            Phi[i - 1] <- Ptr[Phi[i], i]
        }
        
        states <- unique.ss[Phi]
        p$PredictedStructure[k] <- paste(states, collapse="")
    }
    return(p)
}

# Required params:
# Transform E_probs (n x m) into E (simply log)
E <- log(E_probs)
# I think rows are current, cols are next state (need to check viterbi.r for this)
Tr <- log(T_probs)
# I is a numeric of length n, just need to log
I_ <- log(initial_probs)

# We add a new column to both proteins test.tsv and proteins new.tsv,
proteins_test$AminoAcids <- proteins_test$X2
proteins_new$AminoAcids <- proteins_new$X2
proteins_test$PredictedStructure <- NA
proteins_new$PredictedStructure <- NA

proteins_test <- viterbi(E = E, Tr = Tr, I = I_, p = proteins_test)
proteins_new <- viterbi(E = E, Tr = Tr, I = I_, p = proteins_new)

readr::write_tsv(proteins_new, "proteins_new_extended.tsv", col_names = FALSE)
```

\(e\) Estimate confidence intervals for each parameter in I,E and T with bootstrapping. In a single bootstrap run i estimate the probabilities for Ii,Ei and Ti the same as before, but not on the original data set proteins train.tsv, but on the resampled data set. I.e., sample with replacement as many rows from proteins train.tsv as the original data set has. Run a thousand bootstraps and compute the empirical 95% confidence intervals for each single parameter in {Ii}i ,{Ei}i and {Ti}i. 

```{r}

# Create a func to eval. all the params at once:
estimate_params_fast <- function(df) {
  # Assuming X3 is latent var, X2 is the observed
  
  # INITIAL PROBABILITIES (I)
  first_chars <- substr(df$X3, 1, 1)
  I_counts <- table(factor(first_chars, levels = states_vocab))
  I_probs <- as.numeric(I_counts / sum(I_counts))
  names(I_probs) <- states_vocab
  
  # Unlist all sequences simultaneously
  all_states <- unlist(strsplit(df$X3, ""))
  all_obs <- unlist(strsplit(df$X2, ""))
  
  # EMISSION PROBABILITIES (E)
  E_counts <- table(factor(all_states, levels = states_vocab), 
                    factor(all_obs, levels = amino_vocab))
  E_probs <- (E_counts) / rowSums(E_counts)
  
  # TRANSITION PROBABILITIES (T)
  # Drop the last char for "current", drop the first char for "next"
  ss_list <- strsplit(df$X3, "")
  curr_state <- unlist(lapply(ss_list, function(x) x[-length(x)]))
  next_state <- unlist(lapply(ss_list, function(x) x[-1]))
  
  T_counts <- table(factor(curr_state, levels = states_vocab), 
                    factor(next_state, levels = states_vocab))
  T_probs <- (T_counts) / rowSums(T_counts)
  
  return(list(I = I_probs, E = as.matrix(E_probs), Tr = as.matrix(T_probs)))
}

B <- 1000    # Number of bootstraps
n_rows <- nrow(proteins_train)
# Pre-allocate containers to hold the 1000 different versions of our parameters
# I is a vector, so 1000 of them makes a matrix
boot_I <- matrix(0, nrow = B, ncol = length(states_vocab))

# E and T are matrices, so 1000 of them makes a 3D Array!
boot_E <- array(0, dim = c(length(states_vocab), length(amino_vocab), B))
boot_T <- array(0, dim = c(length(states_vocab), length(states_vocab), B))

for (b in 1:B) {
  
  # Sample row indices WITH replacement
  resampled_indices <- sample(1:n_rows, size = n_rows, replace = TRUE)
  
  # Create the new fake dataset
  boot_data <- proteins_train[resampled_indices, ]
  
  # Estimate parameters on the fake dataset
  params <- estimate_params_fast(boot_data)
  
  # Store the results in our containers
  boot_I[b, ]   <- params$I
  boot_E[,,b] <- params$E
  boot_T[,,b] <- params$Tr
}

# The apply function lets us calculate the 2.5% and 97.5% quantiles 
# across the 'bootstrap' dimension for every single parameter simultaneously.

# For Initial Probabilities (margin 2 = columns/states)
CI_I <- apply(boot_I, 2, quantile, probs = c(0.025, 0.975), na.rm = TRUE)

# For Emission Matrix (margins 1 and 2 = row/col combinations)
CI_E_lower <- apply(boot_E, c(1, 2), quantile, probs = 0.025, na.rm = TRUE)
CI_E_upper <- apply(boot_E, c(1, 2), quantile, probs = 0.975, na.rm = TRUE)

# For Transition Matrix (margins 1 and 2 = row/col combinations)
CI_T_lower <- apply(boot_T, c(1, 2), quantile, probs = 0.025, na.rm = TRUE)
CI_T_upper <- apply(boot_T, c(1, 2), quantile, probs = 0.975, na.rm = TRUE)

# Add row/col names back for readability
colnames(CI_I) <- states_vocab
dimnames(CI_E_lower) <- list(states_vocab, amino_vocab)
dimnames(CI_E_upper) <- list(states_vocab, amino_vocab)
dimnames(CI_T_lower) <- list(states_vocab, states_vocab)
dimnames(CI_T_upper) <- list(states_vocab, states_vocab)
```

\(f\) Use the following measure to compute the accuracy of the predicted secondary structure P = (pi) for the data.frame of proteins test.tsv given the real secondary structure S = (si): a(P,S) = 1 L i 1 if pi = si 0 if pi ̸= si with sequence length L. Compute the accuracy for every protein in your data.frame and store the accuracies in a vector. What is the accuracy of the Viterbi algorithm over all sequences (i.e. call summary on the vector of accuracies).

```{r}
# Define the accuracy function
compute_accuracy <- function(pred_seq, true_seq) {
  p_vec <- strsplit(pred_seq, "")[[1]]
  t_vec <- strsplit(true_seq, "")[[1]]
  
  # Accuracy is the proportion of matching characters
  sum(p_vec == t_vec) / length(t_vec)
}

# Calculate accuracy for every sequence in the test set
# (Assuming X3 contains the true secondary structure)
accuracies <- mapply(compute_accuracy, 
                     proteins_test$PredictedStructure, 
                     proteins_test$X3)

# Summarize the accuracy over all sequences
summary(accuracies)
```