---
title: "SMCB_Homework02"
author: "Yuliyan Slavchev, Aurimas Greicius, Dario Franzen"
date: 25 February 2026
format: pdf
---

# Problem 6

## a) Fill the corresponding probabilities (in $\log_{10}$) in the table below

0.  Initialization: 
$$
\begin{aligned}
v_c(0) &= -0.4 \\
v_{nc}(0) &= -0.2
\end{aligned}
$$

1.  Iteration (T) 
$$
\begin{aligned}
v_c(1)  &= E_{c,T} + v_c(0) \\
        &= -0.4 + (-0.4) \\
        &= -0.8 \quad\quad(\text{Originated from } v_c) \\
\\
v_{nc}(1) &= E_{nc,T} + v_{nc}(0) \\
          &= -1.0 + (-0.2) \\
          &= -1.2 \quad\quad(\text{Originated from } v_{nc})
\end{aligned}
$$

2.  Iteration (A) 
$$
\begin{aligned}
v_c(2)  &= E_{c,A} + \max(v_c(1) + T_{c,c}, v_{nc}(1) + T_{nc,c}) \\
        &= E_{c,A} + \max(-0.8 + (-0.5), -1.2 + (-0.2)) \\
        &= E_{c,A} + \max(-1.3, -1.4) \\
        &= -0.7 + (-1.3) \\
        &= -2.0 \quad\quad(\text{Originated from } v_c) \\
\\
v_{nc}(2) &= E_{nc,A} + \max(v_{nc}(1) + T_{nc,nc}, v_c(1) + T_{c,nc}) \\
          &= E_{nc,A} + \max(-1.2 + (-0.4), -0.8 + (-0.15)) \\
          &= E_{nc,A} + \max(-1.6, -0.95) \\
          &= -0.7 + (-0.95) \\
          &= -1.65 \quad\quad(\text{Originated from } v_c)
\end{aligned}
$$

3.  Iteration (A)
$$
\begin{aligned}
v_c(3)  &= E_{c,A} + \max(v_c(2) + T_{c,c}, v_{nc}(2) + T_{nc,c}) \\
        &= E_{c,A} + \max(-2 + (-0.5), -1.65 + (-0.2)) \\
        &= E_{c,A} + \max(-2.5, -1.85) \\
        &= -0.7 + (-1.85) \\
        &= -2.55 \quad\quad(\text{Originated from } v_{nc}) \\
\\
v_{nc}(3) &= E_{nc,A} + \max(v_{nc}(2) + T_{nc,nc}, v_c(2) + T_{c,nc}) \\
          &= E_{nc,A} + \max(-1.65 + (-0.4), -2 + (-0.15)) \\
          &= E_{nc,A} + \max(-2.05, -2.15) \\
          &= -0.7 + (-2.05) \\
          &= -2.75 \quad\quad(\text{Originated from } v_{nc})
\end{aligned}
$$

4.  Iteration (C)

$$
\begin{aligned}
v_c(4)  &= E_{c,C} + \max(v_c(3) + T_{c,c}, v_{nc}(3) + T_{nc,c}) \\
        &= E_{c,C} + \max(-2.55 + (-0.5), -2.75 + (-0.2)) \\
        &= E_{c,C} + \max(-3.05, -2.95) \\
        &= -0.5 + (-2.95) \\
        &= -3.45 \quad\quad(\text{Originated from } v_{nc}) \\
\\
v_{nc}(4) &= E_{nc,C} + \max(v_{nc}(3) + T_{nc,nc}, v_c(3) + T_{c,nc}) \\
          &= E_{nc,C} + \max(-2.75 + (-0.4), -2.55 + (-0.15)) \\
          &= E_{nc,C} + \max(-3.15, -2.70) \\
          &= -0.2 + (-2.70) \\
          &= -2.9 \quad\quad(\text{Originated from } v_c)
\end{aligned}
$$

+------------+------------+--------------+-----------------+-----------------+
|            | T          | A            | A               | C               |
+============+============+==============+=================+=================+
| C-Region   | -0.8       | -2.0         | -2.55           | -3.45           |
|            |            |              |                 |                 |
|            |            | (From $v_c$) | (From $v_{nc}$) | (From $v_{nc}$) |
+------------+------------+--------------+-----------------+-----------------+
| NC-Region  | -1.2       | -1.65        | -2.75           | -2.9            |
|            |            |              |                 |                 |
|            |            | (From $v_c$) | (From $v_{nc}$) | (From $v_c$)    |
+------------+------------+--------------+-----------------+-----------------+


## b) Provide the most likely hidden path for the given sequence (TAAC).

Final sequence with highest probability: C-NC-C-NC

# Problem 7

```{r}
# Import packages
library(tidyr)
```


## a) Read `proteins_train.tsv`, `proteins_test.tsv` and `proteins_new.tsv` into memory and store each in a `data.frame`.

```{r}
proteins_train <- read.delim("data/proteins_train.tsv", header = F, 
                             stringsAsFactors = FALSE)
proteins_test <- read.delim("data/proteins_test.tsv", header = F,
                            stringsAsFactors = FALSE)
proteins_new <- read.delim("data/proteins_new.tsv", header = F,
                           stringsAsFactors = FALSE)
```

## b) Estimate the vector of initial state probabilities $I$, the matrix of transition probabilities $T$ and the matrix for emission probabilities $E$ by using maximum likelihood, i.e., by counting how often each transition and each emission happened and how often each secondary structure was found at the beginning of a protein, respectively.

```{r}
observed_state_space <- c("A", "C", "D", "E", "F", "G", "H", "I", "K", "L", 
                          "M", "N", "P", "Q", "R", "S", "T", "U", "V", "W", 
                          "X", "Y")
hidden_state_space <- c("B", "C", "E", "G", "H", "I", "S", "T")

# Initial state probabilities I
init_prob <- function(hid_state, data) {
  # Collect first letters
  first_letters <- substr(data[,3], 1, 1)
  counts <- table(factor(first_letters, levels = hid_state))
  
  # Add pseudocount and normalize
  initial_state_prob <- counts + 1
  initial_state_prob <- initial_state_prob / sum(initial_state_prob)
  
  return(as.numeric(initial_state_prob))
}

initial_state_prob <- init_prob(hidden_state_space, proteins_train)

```

```{r}
# Transition probabilities T
tran_prob <- function(hid_state, data) {
  # Split strings into lists of characters
  hid_list <- strsplit(data[,3], "")
  
  # Go through each sequence and drop the last letter
  from_chars <- unlist(lapply(hid_list, function(x) x[-length(x)]))
  
  # Go through each sequence and drop the first letter
  to_chars <- unlist(lapply(hid_list, function(x) x[-1]))
  
  # Create the transition matrix
  counts <- table(
    factor(from_chars, levels = hid_state),
    factor(to_chars, levels = hid_state)
  )
  
  # Add pseudocounts and normalize
  transition_prob <- counts + 1
  transition_prob <- transition_prob / rowSums(transition_prob)
  
  return(matrix(transition_prob, nrow = length(hid_state), 
                dimnames = list(hid_state, hid_state)))
}

transition_prob <- tran_prob(hidden_state_space, proteins_train)


```

```{r}
# Emission probabilities E
emis_prob <- function(hid_state, obs_state, data) {
  all_obs_chars <- unlist(strsplit(data[,2], ""))
  all_hid_chars <- unlist(strsplit(data[,3], ""))
  
  # Map the two vectors
  counts <- table(
    factor(all_hid_chars, levels = hid_state),
    factor(all_obs_chars, levels = obs_state)
  )
  
  # Add pseudocounts and normalize
  emission_prob <- counts + 1
  emission_prob <- emission_prob / rowSums(emission_prob)
  
  return(matrix(emission_prob, nrow = length(hid_state), 
                dimnames = list(hid_state, obs_state)))
}

emission_prob <- emis_prob(hidden_state_space, observed_state_space, 
                           proteins_train)

```


## c)  Estimate the stationary distribution $\pi$ of the Markov chain

```{r}
# Get eigenvectors of T transposed
eig <- eigen(t(transition_prob))
pi <- Re(eig$vectors[, 1])

pi <- pi / sum(pi)

# Check if the same
all.equal(as.numeric(t(pi)), as.numeric(t(pi) %*% transition_prob))

pi
```

## d) Predict secondary structure of protein_test and protein_new using viterbi

```{r}
source("code/viterbi.r")

# Transform all probabilities into log space to be used in the function
log_E <- log(emission_prob)
log_Tr <- log(transition_prob)
log_I <- log(initial_state_prob)

# First prediction on protein_test
proteins_test_pred <- proteins_test
names(proteins_test_pred) <- c("ProteinID", "AminoAcids", "SecondStruct")

proteins_test_pred <- viterbi(
  E = log_E,
  Tr = log_Tr,
  I = log_I,
  p = proteins_test_pred
)

# Second prediction on protein_new
proteins_new_pred <- proteins_new
names(proteins_new_pred) <- c("ProteinID", "AminoAcids")

proteins_new_pred <- viterbi(
  E = log_E,
  Tr = log_Tr,
  I = log_I,
  p = proteins_new_pred
)

# Write protein_new_pred as file
write.table(proteins_new_pred, file="proteins_new_pred.tsv", 
            quote=FALSE, sep="\t", col.names = FALSE, row.names = FALSE)

```

## e) Estimate confidence intervals for each parameter in $I$, $E$ and $T$ with bootstrapping.

```{r}
# Number of rows of proteins_train
n_rows <- nrow(proteins_train)
n_hidden <- length(hidden_state_space)
n_observed <- length(observed_state_space)

# Container to hold the calculated probs
boot_I <- matrix(NA, nrow = 1000, ncol = n_hidden)
colnames(boot_I) <- hidden_state_space

boot_E <- array(NA, dim = c(n_hidden, n_observed, 1000))
dimnames(boot_E) <- list(hidden_state_space, observed_state_space, NULL)

boot_T <- array(NA, dim = c(n_hidden, n_hidden, 1000))
dimnames(boot_T) <- list(hidden_state_space, hidden_state_space, NULL)

# Calculate probabilites 1000 times
for (i in 1:1000) {
  boot_indices <- sample(1:n_rows, size = n_rows, replace = TRUE)
  proteins_train_boot <- proteins_train[boot_indices, ]
  
  # Call functions
  boot_init <- init_prob(hidden_state_space, proteins_train_boot)
  boot_emis <- emis_prob(hidden_state_space, observed_state_space, 
                         proteins_train_boot)
  boot_tran <- tran_prob(hidden_state_space, proteins_train_boot)
  
  # Save values
  boot_I[i, ] <- boot_init
  boot_E[, , i] <- boot_emis
  boot_T[, , i] <- boot_tran
}
```

```{r}
# Get the 95% CI for every parameter
CI_I <- apply(boot_I, 2, function(x) quantile(x, probs = c(0.025, 0.975)))
CI_E <- apply(boot_E, c(1, 2), function(x) quantile(x, 
                                                    probs = c(0.025, 0.975)))
CI_T <- apply(boot_T, c(1, 2), function(x) quantile(x, 
                                                    probs = c(0.025, 0.975)))
```


## f) Compute the accuracy of the secondary structure predicitons

```{r}
# Create a vector to hold accuracy values
accuracy <- c(rep(0, nrow(proteins_test_pred)))
names(accuracy) <- proteins_test_pred$ProteinID

# Iterate over rows in protein_test
for (i in 1:nrow(proteins_test_pred)) {
  current_secStruc <- proteins_test_pred[i,3]
  current_pred <- proteins_test_pred[i,4]
  
  # Iterate over chars in the current secondary structure and prediction
  for (j in 1:nchar(current_secStruc)) {
    if (substr(current_secStruc, j, j) == substr(current_pred, j, j)) {
      accuracy[proteins_test_pred[i,1]] <- accuracy[
        proteins_test_pred[i,1]] + 1
    }
  }
  
  accuracy[proteins_test_pred[i,1]] <- accuracy[
    proteins_test_pred[i,1]] / nchar(current_secStruc)
}

summary(accuracy)

```

