---
title: "Homework 7: Variational inference"
author:
  - Bernhard Buss
  - Dario Franzen
  - Aurimas Greicius
  - Timothy Lauren Nijhuis
  - Carlo Rossetto
  - Yuliyan Slavchev
  - Görkem Kadir Solun
format: pdf
editor: visual
---

```{r}
set.seed(42)
```

## Problem 17: Variational inference

### Part a: Write down the full joint probability & log-probability of the data and latent variables.

Let $x=(x_1,\dots,x_N)$, $z=(z_1,\dots,z_N)$ with $z_i\in\{1,\dots,K\}$, mixing proportions $\pi=(\pi_1,\dots,\pi_K)$, and component parameters $p=(p_1,\dots,p_K)$. The model is:

$$
p(x_i\mid z_i=k,p_k)=\binom{x_i+r-1}{x_i}p_k^r(1-p_k)^{x_i},\qquad
p(z_i=k\mid\pi)=\pi_k,
$$

with priors

$$
\pi\sim\mathrm{Dirichlet}(\alpha_{0,1},\dots,\alpha_{0,K}),\qquad
p_k\sim\mathrm{Beta}(a_0,b_0).
$$

Hence the full joint factorization is

$$
p(x,z,\pi,p)
=p(\pi)\prod_{k=1}^K p(p_k)\prod_{i=1}^N p(z_i\mid\pi)\,p(x_i\mid z_i,p).
$$

Using one-hot indicators $\mathbb{1}[z_i=k]$:

$$
p(x,z,\pi,p)
=\frac{1}{B(\alpha_0)}\prod_{k=1}^K \pi_k^{\alpha_{0,k}-1}
\prod_{k=1}^K\frac{1}{B(a_0,b_0)}p_k^{a_0-1}(1-p_k)^{b_0-1}
\prod_{i=1}^N\prod_{k=1}^K\left[\pi_k\binom{x_i+r-1}{x_i}p_k^r(1-p_k)^{x_i}\right]^{\mathbb{1}[z_i=k]}.
$$

Therefore, the log-joint is

$$
\begin{aligned}
\log p(x,z,\pi,p)
&= -\log B(\alpha_0) + \sum_{k=1}^K(\alpha_{0,k}-1)\log\pi_k \\
&\quad + \sum_{k=1}^K\Big[-\log B(a_0,b_0) + (a_0-1)\log p_k + (b_0-1)\log(1-p_k)\Big] \\
&\quad + \sum_{i=1}^N\sum_{k=1}^K\mathbb{1}[z_i=k]\Big[\log\pi_k + \log\binom{x_i+r-1}{x_i} + r\log p_k + x_i\log(1-p_k)\Big].
\end{aligned}
$$

### Part b: Derive the variational update for Q(zi)

Use mean-field:

$$
Q(\pi,p,z)=Q(\pi)\prod_{k=1}^KQ(p_k)\prod_{i=1}^NQ(z_i),
$$

with $Q(\pi)=\mathrm{Dirichlet}(\alpha)$ and $Q(p_k)=\mathrm{Beta}(a_k,b_k)$.

For $Q(z_i)$:

$$
\log Q^*(z_i=k)\propto \mathbb{E}_{Q(\pi)Q(p_k)}\big[\log\pi_k+r\log p_k+x_i\log(1-p_k)\big].
$$

Using Dirichlet/Beta expectations,

$$
\mathbb{E}[\log\pi_k]=\psi(\alpha_k)-\psi\!\left(\sum_{j=1}^K\alpha_j\right),
$$

$$
\mathbb{E}[\log p_k]=\psi(a_k)-\psi(a_k+b_k),\qquad
\mathbb{E}[\log(1-p_k)]=\psi(b_k)-\psi(a_k+b_k),
$$

so responsibilities are

$$
\rho_{ik}:=Q(z_i=k)\propto
\exp\!\left\{
\psi(\alpha_k)-\psi\!\left(\sum_{j=1}^K\alpha_j\right)
+r\,[\psi(a_k)-\psi(a_k+b_k)]
+x_i\,[\psi(b_k)-\psi(a_k+b_k)]
\right\},
$$

then normalize:

$$
Q(z_i=k)=\frac{\rho_{ik}}{\sum_{\ell=1}^K\rho_{i\ell}},\qquad
\sum_{k=1}^KQ(z_i=k)=1.
$$

### Part c: Derive the variational update for Q(π)

Collect terms involving $\pi$:

$$
\log Q^*(\pi)\propto \sum_{k=1}^K\left(\alpha_{0,k}-1+\sum_{i=1}^N\mathbb{E}[\mathbb{1}[z_i=k]]\right)\log\pi_k.
$$

Since $\mathbb{E}[\mathbb{1}[z_i=k]]=Q(z_i=k)$, we get

$$
Q(\pi)=\mathrm{Dirichlet}(\alpha_1,\dots,\alpha_K),
$$

$$
\alpha_k=\alpha_{0,k}+\sum_{i=1}^N Q(z_i=k).
$$

### Part d: Derive the variational update for Q(pk)

Collect terms involving one $p_k$:

$$
\log Q^*(p_k)\propto
\left(a_0-1+r\sum_{i=1}^N\mathbb{E}[\mathbb{1}[z_i=k]]\right)\log p_k
+\left(b_0-1+\sum_{i=1}^N x_i\,\mathbb{E}[\mathbb{1}[z_i=k]]\right)\log(1-p_k).
$$

Thus

$$
Q(p_k)=\mathrm{Beta}(a_k,b_k),
$$

$$
a_k=a_0+r\sum_{i=1}^NQ(z_i=k),\qquad
b_k=b_0+\sum_{i=1}^NQ(z_i=k)x_i.
$$

These are the CAVI updates used iteratively with the $Q(z_i)$ update until convergence of the ELBO (or parameter change threshold).

## Problem 18: R-Implementation of variational inference updating equations for a Negative Binomial mixture of cell populations

### Part a: Implement the above derived variational inference procedure to estimate λk and π for the data in vi.csv.

We use the parameterization

$$
X\mid z_i=k\sim\mathrm{NegBin}(r,p_k),\qquad
\mathbb{E}[X\mid z_i=k]=\lambda_k=\frac{r(1-p_k)}{p_k}.
$$

```{r}
# Data loading helper
load_vi_counts <- function(preferred_col = NULL) {
  csv_candidates <- c("vi.csv", "hw7/vi.csv")
  csv_path <- csv_candidates[file.exists(csv_candidates)][1]

  if (!is.na(csv_path)) {
    dat <- read.csv(csv_path)
  } else {
    stop("vi.csv not found (checked local folder and hw7/).")
  }

  # Select count column.
  if (!is.null(preferred_col) && preferred_col %in% names(dat)) {
    x <- dat[[preferred_col]]
  } else {
    numeric_cols <- vapply(dat, is.numeric, logical(1))
    if (!any(numeric_cols)) {
      stop("No numeric column found in input file.")
    }

    idx <- which(numeric_cols)
    nms <- names(dat)[idx]

    # Drop index-like columns if possible (typical in exported spreadsheets).
    index_like <- grepl("^(\\.\\.\\.[0-9]+|id|index|cell(_?id)?)$", tolower(nms))
    if (any(!index_like)) {
      idx <- idx[!index_like]
      nms <- nms[!index_like]
    }

    # Prefer biologically meaningful names when available.
    preferred_name <- grepl("atp|count|expr|expression|gene", tolower(nms))
    if (any(preferred_name)) {
      idx <- idx[preferred_name]
    }

    # Final fallback: choose column with smallest unique-ratio (less likely to be row index).
    unique_ratio <- vapply(
      dat[idx],
      function(col) length(unique(col[!is.na(col)])) / sum(!is.na(col)),
      numeric(1)
    )
    x <- dat[[idx[which.min(unique_ratio)]]]
  }

  if (any(is.na(x))) {
    x <- x[!is.na(x)]
  }

  x <- as.integer(round(x))
  if (any(x < 0)) {
    stop("Counts must be non-negative.")
  }

  x
}

# CAVI for NB mixture
run_nb_vi <- function(x,
                      K = 2,
                      r = 10,
                      a0 = 1,
                      b0 = 1,
                      alpha0 = 1,
                      max_iter = 1000,
                      tol = 1e-8,
                      seed = 42) {
  set.seed(seed)

  N <- length(x)
  alpha0_vec <- rep(alpha0, K)

  # Random uniform initialization for q(z_i = k), row-normalized.
  resp <- matrix(runif(N * K, min = 0, max = 1), nrow = N, ncol = K)
  resp <- resp / rowSums(resp)

  alpha <- alpha0_vec + colSums(resp)
  a <- a0 + r * colSums(resp)
  b <- b0 + colSums(resp * x)
  converged <- FALSE

  for (iter in seq_len(max_iter)) {
    resp_old <- resp

    # Update q(z_i): use log-space for stability.
    elog_pi <- digamma(alpha) - digamma(sum(alpha))
    elog_p <- digamma(a) - digamma(a + b)
    elog_1mp <- digamma(b) - digamma(a + b)

    for (i in seq_len(N)) {
      log_rho <- elog_pi + r * elog_p + x[i] * elog_1mp
      log_rho <- log_rho - max(log_rho)
      rho <- exp(log_rho)
      resp[i, ] <- rho / sum(rho)
    }

    # Update q(pi).
    alpha <- alpha0_vec + colSums(resp)

    # Update q(p_k).
    a <- a0 + r * colSums(resp)
    b <- b0 + colSums(resp * x)

    # Convergence check on responsibilities.
    delta <- max(abs(resp - resp_old))
    if (delta < tol) {
      converged <- TRUE
      break
    }
  }

  # Posterior means as point estimates.
  pi_hat <- alpha / sum(alpha)
  p_hat <- a / (a + b)
  lambda_hat <- r * (1 - p_hat) / p_hat

  list(
    r = r,
    iterations = iter,
    converged = converged,
    alpha = alpha,
    a = a,
    b = b,
    resp = resp,
    pi_hat = pi_hat,
    p_hat = p_hat,
    lambda_hat = lambda_hat
  )
}

# Run for r in {10, 100, 1000}
# with vague priors a0=b0=alpha0=1
x <- load_vi_counts()
r_values <- c(10, 100, 1000)

vi_results <- lapply(
  r_values,
  function(r) {
    run_nb_vi(
      x = x,
      K = 2,
      r = r,
      a0 = 1,
      b0 = 1,
      alpha0 = 1,
      max_iter = 1000,
      tol = 1e-8,
      seed = 42
    )
  }
)

# Report p_k, pi_k, lambda_k
summary_table <- do.call(
  rbind,
  lapply(vi_results, function(res) {
    data.frame(
      r = res$r,
      component = seq_along(res$p_hat),
      p_k = res$p_hat,
      lambda_k = res$lambda_hat,
      pi_k = res$pi_hat,
      iterations = res$iterations,
      converged = res$converged
    )
  })
)

summary_table
```
