---
title: "Homework 7: Variational inference"
author:
  - Bernhard Buss
  - Dario Franzen
  - Aurimas Greicius
  - Timothy Lauren Nijhuis
  - Carlo Rossetto
  - Yuliyan Slavchev
  - Görkem Kadir Solun
format: pdf
editor: visual
---

```{r}
set.seed(42)
```

## Problem 17: Derivation variational inference updating equation for anegative binomial mixture of cell populations

### Part a: Write down the full joint probability & log-probability of the data and latent variables.

Let $x=(x_1,\dots,x_N)$, $z=(z_1,\dots,z_N)$ with $z_i\in\{1,\dots,K\}$, mixing proportions $\pi=(\pi_1,\dots,\pi_K)$, and component parameters $p=(p_1,\dots,p_K)$. The model is:

$$
p(x_i\mid z_i=k,p_k)=\binom{x_i+r-1}{x_i}p_k^r(1-p_k)^{x_i},\qquad
p(z_i=k\mid\pi)=\pi_k,
$$

with priors

$$
\pi\sim\mathrm{Dirichlet}(\alpha_{0,1},\dots,\alpha_{0,K}),\qquad
p_k\sim\mathrm{Beta}(a_0,b_0).
$$

Hence the full joint factorization is

$$
p(x,z,\pi,p)
=p(\pi)\prod_{k=1}^K p(p_k)\prod_{i=1}^N p(z_i\mid\pi)\,p(x_i\mid z_i,p).
$$

Using one-hot indicators $\mathbb{1}[z_i=k]$:

$$
p(x,z,\pi,p)
=\frac{1}{B(\alpha_0)}\prod_{k=1}^K \pi_k^{\alpha_{0,k}-1}
\prod_{k=1}^K\frac{1}{B(a_0,b_0)}p_k^{a_0-1}(1-p_k)^{b_0-1}
\prod_{i=1}^N\prod_{k=1}^K\left[\pi_k\binom{x_i+r-1}{x_i}p_k^r(1-p_k)^{x_i}\right]^{\mathbb{1}[z_i=k]}.
$$

Therefore, the log-joint is

$$
\begin{aligned}
\log p(x,z,\pi,p)
&= -\log B(\alpha_0) + \sum_{k=1}^K(\alpha_{0,k}-1)\log\pi_k \\
&\quad + \sum_{k=1}^K\Big[-\log B(a_0,b_0) + (a_0-1)\log p_k + (b_0-1)\log(1-p_k)\Big] \\
&\quad + \sum_{i=1}^N\sum_{k=1}^K\mathbb{1}[z_i=k]\Big[\log\pi_k + \log\binom{x_i+r-1}{x_i} + r\log p_k + x_i\log(1-p_k)\Big].
\end{aligned}
$$

### Part b: Derive the variational update for Q(z_i)

Use mean-field:

$$
Q(\pi,p,z)=Q(\pi)\prod_{k=1}^KQ(p_k)\prod_{i=1}^NQ(z_i),
$$

with $Q(\pi)=\mathrm{Dirichlet}(\alpha)$ and $Q(p_k)=\mathrm{Beta}(a_k,b_k)$.

For $Q(z_i)$:

$$
\log Q^*(z_i=k)\propto \mathbb{E}_{Q(\pi)Q(p_k)}\big[\log\pi_k+r\log p_k+x_i\log(1-p_k)\big].
$$

Using Dirichlet/Beta expectations,

$$
\mathbb{E}[\log\pi_k]=\psi(\alpha_k)-\psi\!\left(\sum_{j=1}^K\alpha_j\right),
$$

$$
\mathbb{E}[\log p_k]=\psi(a_k)-\psi(a_k+b_k),\qquad
\mathbb{E}[\log(1-p_k)]=\psi(b_k)-\psi(a_k+b_k),
$$

so responsibilities are

$$
\rho_{ik}:=Q(z_i=k)\propto
\exp\!\left\{
\psi(\alpha_k)-\psi\!\left(\sum_{j=1}^K\alpha_j\right)
+r\,[\psi(a_k)-\psi(a_k+b_k)]
+x_i\,[\psi(b_k)-\psi(a_k+b_k)]
\right\},
$$

then normalize:

$$
Q(z_i=k)=\frac{\rho_{ik}}{\sum_{\ell=1}^K\rho_{i\ell}},\qquad
\sum_{k=1}^KQ(z_i=k)=1.
$$

### Part c: Derive the variational update for Q\[π(pi)\]

Collect terms involving $\pi$:

$$
\log Q^*(\pi)\propto \sum_{k=1}^K\left(\alpha_{0,k}-1+\sum_{i=1}^N\mathbb{E}[\mathbb{1}[z_i=k]]\right)\log\pi_k.
$$

Since $\mathbb{E}[\mathbb{1}[z_i=k]]=Q(z_i=k)$, we get

$$
Q(\pi)=\mathrm{Dirichlet}(\alpha_1,\dots,\alpha_K),
$$

$$
\alpha_k=\alpha_{0,k}+\sum_{i=1}^N Q(z_i=k).
$$

### Part d: Derive the variational update for Q(p_k)

Collect terms involving one $p_k$:

$$
\log Q^*(p_k)\propto
\left(a_0-1+r\sum_{i=1}^N\mathbb{E}[\mathbb{1}[z_i=k]]\right)\log p_k
+\left(b_0-1+\sum_{i=1}^N x_i\,\mathbb{E}[\mathbb{1}[z_i=k]]\right)\log(1-p_k).
$$

Thus

$$
Q(p_k)=\mathrm{Beta}(a_k,b_k),
$$

$$
a_k=a_0+r\sum_{i=1}^NQ(z_i=k),\qquad
b_k=b_0+\sum_{i=1}^NQ(z_i=k)x_i.
$$

These are the CAVI updates used iteratively with the $Q(z_i)$ update until convergence of the ELBO (or parameter change threshold).

## Problem 18: R-Implementation of variational inference updating equations for a Negative Binomial mixture of cell populations

### Part a: Implement the above derived variational inference procedure to estimate λk and π for the data in vi.csv.

We use the parameterization

$$
X\mid z_i=k\sim\mathrm{NegBin}(r,p_k),\qquad
\mathbb{E}[X\mid z_i=k]=\lambda_k=\frac{r(1-p_k)}{p_k}.
$$

```{r}
# Load data: second column contains ATP6V0E1 counts
dat <- read.csv("vi.csv")
x <- dat[[2]]
N <- length(x)
cat("N =", N, "| mean =", round(mean(x), 3), "| var =", round(var(x), 3), "\n")
```

```{r}
# ELBO computation for the NB mixture CAVI
compute_elbo <- function(x, resp, alpha, a, b, alpha0_vec, a0, b0, r, K) {
  N <- length(x)

  # E[log p(pi)] - E[log q(pi)]
  alpha_sum <- sum(alpha)
  alpha0_sum <- sum(alpha0_vec)
  elbo <- lgamma(alpha0_sum) - sum(lgamma(alpha0_vec)) +
    sum((alpha0_vec - 1) * (digamma(alpha) - digamma(alpha_sum))) -
    (lgamma(alpha_sum) - sum(lgamma(alpha)))

  # E[log p(p_k)] - E[log q(p_k)]
  for (k in seq_len(K)) {
    e_log_pk <- digamma(a[k]) - digamma(a[k] + b[k])
    e_log_1mpk <- digamma(b[k]) - digamma(a[k] + b[k])
    elbo <- elbo +
      (a0 - 1) * e_log_pk + (b0 - 1) * e_log_1mpk -
      lgamma(a0) - lgamma(b0) + lgamma(a0 + b0) -
      ((a[k] - 1) * e_log_pk + (b[k] - 1) * e_log_1mpk -
        lgamma(a[k]) - lgamma(b[k]) + lgamma(a[k] + b[k]))
  }

  # E[log p(z|pi)] + E[log p(x|z,p)] - E[log q(z)]
  elog_pi <- digamma(alpha) - digamma(alpha_sum)
  elog_p <- digamma(a) - digamma(a + b)
  elog_1mp <- digamma(b) - digamma(a + b)

  for (k in seq_len(K)) {
    Nk <- resp[, k]
    elbo <- elbo + sum(Nk * elog_pi[k])
    elbo <- elbo + sum(Nk * (lchoose(x + r - 1, x) + r * elog_p[k] + x * elog_1mp[k]))
    safe_log <- ifelse(Nk > 1e-300, log(Nk), 0)
    elbo <- elbo - sum(Nk * safe_log)
  }

  elbo
}

# CAVI for NB mixture (vectorized)
run_nb_vi <- function(x, K = 2, r = 10, a0 = 1, b0 = 1, alpha0 = 1,
                      max_iter = 10000, tol = 1e-6, seed = 42) {
  set.seed(seed)
  N <- length(x)
  alpha0_vec <- rep(alpha0, K)

  # Random uniform initialization for q(z_i = k), row-normalized.
  resp <- matrix(runif(N * K), nrow = N, ncol = K)
  resp <- resp / rowSums(resp)

  alpha <- alpha0_vec + colSums(resp)
  a <- a0 + r * colSums(resp)
  b <- b0 + colSums(resp * x)
  converged <- FALSE
  elbo_trace <- numeric(max_iter)

  for (iter in seq_len(max_iter)) {
    resp_old <- resp

    # Update q(z_i) — vectorized over all observations.
    elog_pi <- digamma(alpha) - digamma(sum(alpha))
    elog_p <- digamma(a) - digamma(a + b)
    elog_1mp <- digamma(b) - digamma(a + b)

    # log_rho is N x K: each row = elog_pi + r*elog_p + x_i*elog_1mp
    log_rho <- outer(rep(1, N), elog_pi + r * elog_p) +
      outer(x, elog_1mp)
    log_rho <- log_rho - apply(log_rho, 1, max)
    resp <- exp(log_rho)
    resp <- resp / rowSums(resp)

    # Update q(pi).
    alpha <- alpha0_vec + colSums(resp)

    # Update q(p_k).
    a <- a0 + r * colSums(resp)
    b <- b0 + colSums(resp * x)

    # ELBO tracking.
    elbo_trace[iter] <- compute_elbo(x, resp, alpha, a, b, alpha0_vec, a0, b0, r, K)

    # Convergence check on responsibilities.
    delta <- max(abs(resp - resp_old))
    if (delta < tol) {
      converged <- TRUE
      break
    }
  }

  pi_hat <- alpha / sum(alpha)
  p_hat <- a / (a + b)
  lambda_hat <- r * (1 - p_hat) / p_hat

  list(
    r = r, iterations = iter, converged = converged,
    alpha = alpha, a = a, b = b, resp = resp,
    pi_hat = pi_hat, p_hat = p_hat, lambda_hat = lambda_hat,
    elbo_trace = elbo_trace[seq_len(iter)]
  )
}
```

```{r}
# Run for r in {10, 100, 1000} with vague priors a0=b0=alpha0=1
r_values <- c(10, 100, 1000)

# increased number of iterations from proposed 1,000 to 10,000 to ensure convergence
vi_results <- lapply(r_values, function(r) {
  run_nb_vi(x, K = 2, r = r, a0 = 1, b0 = 1, alpha0 = 1,
            max_iter = 10000, tol = 1e-6, seed = 42)
})

# Report p_k, pi_k, lambda_k
summary_table <- do.call(rbind, lapply(vi_results, function(res) {
  data.frame(
    r = res$r,
    component = seq_along(res$p_hat),
    p_k = round(res$p_hat, 6),
    lambda_k = round(res$lambda_hat, 4),
    pi_k = round(res$pi_hat, 4),
    iterations = res$iterations,
    converged = res$converged
  )
}))
rownames(summary_table) <- NULL
summary_table
```
From the table above it is evident that the choice of $r$ is important. At $r=10$ (high overdispersion), the model explains the data by isolating a very small, near-zero expression group ($\pi_2 \approx 7\%$, $\lambda_2 \approx 0.12$). However, as $r$ increases to $100$ and $1000$, the parameter estimates stabilize into two well-proportioned, distinct components. The similarity between the $r=100$ and $r=1000$ estimates suggests the model is highly confident in this specific separation.

```{r}
# ELBO convergence plots
par(mfrow = c(1, 3))
for (res in vi_results) {
  plot(res$elbo_trace, type = "l", col = "steelblue",
       xlab = "Iteration", ylab = "ELBO",
       main = paste0("r = ", res$r))
}
```

The CAVI algorithm formally converges (converged = TRUE) for all values of the dispersion parameter $r$. The ELBO plots above confirm this success, showing a smooth, monotonic increase that plateaus, indicating the mean-field approximation has reached a stable local optimum. Interestingly, as $r$ increases, the algorithm converges faster: around 6000 iterations at $r=10$ to around 4000 iterations at $r=1000$.

```{r}
# Histogram of data with fitted NB mixture densities
par(mfrow = c(1, 3))
x_vals <- 0:max(x)
for (res in vi_results) {
  hist(x, breaks = seq(-0.5, max(x) + 0.5, by = 1), freq = FALSE,
       col = "grey85", border = "white",
       main = paste0("r = ", res$r),
       xlab = "Count", ylab = "Density")
  mix_dens <- numeric(length(x_vals))
  colors <- c("red", "blue")
  for (k in seq_along(res$p_hat)) {
    dens_k <- dnbinom(x_vals, size = res$r, prob = res$p_hat[k])
    lines(x_vals, res$pi_hat[k] * dens_k, col = colors[k], lwd = 2)
    mix_dens <- mix_dens + res$pi_hat[k] * dens_k
  }
  lines(x_vals, mix_dens, col = "black", lwd = 2, lty = 2)
  legend("topright", cex = 0.7,
         legend = c(
           bquote(lambda[1] == .(round(res$lambda_hat[1], 2)) ~ ", " ~
                    pi[1] == .(round(res$pi_hat[1], 2))),
           bquote(lambda[2] == .(round(res$lambda_hat[2], 2)) ~ ", " ~
                    pi[2] == .(round(res$pi_hat[2], 2))),
           "Mixture"),
         col = c("red", "blue", "black"), lwd = 2, lty = c(1, 1, 2))
}
```

The plots above confirm what was observed in the results table. Unlike at $r = 10$, where there is a small, lower expression group (blue line in left plot), with $r=100$ and $r=1000$, the model successfully separates the data into a higher-expression group (red line, $\lambda_1 \approx 1.2$, comprising $\approx 66-69\%$ of cells) and a lower-expression group (blue line, $\lambda_2 \approx 0.4$, comprising $\approx 31-34\%$ of cells). Together, their mixture (dotted black line) matches the total distribution (bars) well. These are likely to be the two biological groups,  possibly the two distinct cell types (B-cells and T-cells) characterized by their differential expression of the ATP6V0E1 gene.
