---
title: "Homework 5: Statistical Phylogenetics"
format: pdf
editor: visual
---

---
title: "Homework 5: Statistical Phylogenetics"
author:
  - Bernhard Buss
  - Dario Franzen
  - Aurimas Greicius
  - Timothy Lauren Nijhuis
  - Carlo Rossetto
  - Yuliyan Slavchev
  - Görkem Kadir Solun
date: "March 2026"
format: pdf
---

## Problem 12: Intuition on time-reversible Markov chains

### 1. Time-reversed process transition probabilities

For a time reversed-process:

$$
\hat T_{kl}(s,t):=Pr\left(\hat X(s+t) = l|\hat X(s) = k\right) = P\left(\hat X(s+t) = l|\hat X(s) = k\right)
$$

$$
= P\left(X(\tau-(s+t)) = l| X(\tau - s) = k\right)
$$ Applying the law of conditional probability: $$ = \frac{P\left(X(\tau-s-t) = l, X(\tau - s) = k\right)}{P( X(\tau - s) = k)} = \frac{P\left(X(\tau-s-t) = l, X(\tau - s) = k\right)}{X(0)}=$$ $$\frac{P\left(X(\tau-s-t) = l, X(\tau - s) = k\right)}{\pi_{k}}=$$The numerator is the probability of reaching the state $l$ at time $\tau-s-t$ (which is earlier than $\tau-s$) times the probability of reaching state $k$ a time $t$ later, given state $l$ was reached beforehand:

$$\frac{P\left(X(\tau-s-t) = l\right) \cdot P\left(X(\tau - s) = k|X(\tau-s-t) = l\right)}{\pi_{k}}=\frac{\pi_{l} \cdot T_{lk}(t)}{\pi_{k}}$$ Therefore: $$\rightarrow \hat T_{kl}(s,t)=\frac{\pi_{l} \cdot T_{lk}(t)}{\pi_{k}}$$

### 2. Original Markov chain reversability

Knowing $$
\hat T_{kl}(s,t) = \frac{\pi_{l}\cdot T_{lk}(t)}{\pi_{k}}
$$ and $$
\hat T_{kl}(s,t) = \hat T_{kl}(t) \rightarrow \hat T_{kl}(t) = \frac{\pi_{l}\cdot T_{lk}(t)}{\pi_{k}}
$$ Furthermore, since $$
\pi_{k}\cdot T_{kl}(t) = \pi_{l}\cdot T_{lk}(t)
$$ for all $k,l \in [K], t\ge 0$, we can combine the two equations to obtain: $$
\hat T_{kl}(t) = \frac{\left(\pi_{k}\cdot T_{kl}(t)\right)}{\pi_{k}} = T_{kl}(t) \rightarrow T_{kl}(t) = \hat T_{kl}(t)
$$

\newpage

## Problem 13: Phylogenetic trees as Bayesian networks

### 1. Summation terms in brute force marginalisation

The joint probability $P(Y,Z|T)$ is as follows:

$$ P(Y,Z|T) = P(Z_{1})\cdot P(Y_{4}|Z_{1})\cdot P(Z_{2}|Z_{1})\cdot P(Y_{3}|Z_{2})\cdot P(Z_{3}|Z_{2})\cdot P(Y_{2}|Z{3})\cdot P(Y_{1}|Z{3})$$

Then, marginalising over $Z$ to obtain $P(Y|T)$: $$ P(Y|T) = \sum_{Z}\left(P(Z_{1})\cdot P(Y_{4}|Z_{1})\cdot P(Z_{2}|Z_{1})\cdot P(Y_{3}|Z_{2})\cdot P(Z_{3}|Z_{2})\cdot P(Y_{2}|Z{3})\cdot P(Y_{1}|Z{3})\right) = $$ $$= \sum_{Z_{1},Z_{2},Z_{3}}\left(P(Z_{1})\cdot P(Y_{4}|Z_{1})\cdot P(Z_{2}|Z_{1})\cdot P(Y_{3}|Z_{2})\cdot P(Z_{3}|Z_{2})\cdot P(Y_{2}|Z{3})\cdot P(Y_{1}|Z{3})\right)$$ Since each hidden variables, $Z_{1}$, $Z_{2}$ and $Z_{3}$, we are marginalising \$ has 4 states ${A,C,G,T}$, there are total of 64 summation terms $$4^{3} = 64$$

### 2. Summation terms after rearranging with Felsenstein's peeling algorithm.

Rearrangement with Felsenstein's peeling algorithm results as follows

$$ P(Y|T) = \sum_{Z_{1},Z_{2},Z_{3}}\left(P(Z_{1})\cdot P(Y_{4}|Z_{1})\cdot P(Z_{2}|Z_{1})\cdot P(Y_{3}|Z_{2})\cdot P(Z_{3}|Z_{2})\cdot P(Y_{2}|Z{3})\cdot P(Y_{1}|Z{3})\right)$$ $$= \sum_{Z_{1}}\left(P(Z_{1})\cdot P(Y_{4}|Z_{1})\sum_{Z_{2}}\left(P(Z_{2}|Z_{1})\cdot P(Y_{3}|Z_{2})\sum_{Z_{3}}\left(P(Z_{3}|Z_{2})\cdot P(Y_{2}|Z{3})\cdot P(Y_{1}|Z{3})\right)\right)\right)$$ The number of summation terms is: $$ (m-1)*4^{2}+4 $$ where $m$ is the number of internal nodes, or latent variables. Given there are three internal nodes, this results to be 36 summation terms: $$ (3-1)*4^{2}+4 = 36 $$ Significantly less compared to the naive 'brute force' marginalisation. \newpage

## Problem 14 (data analysis): Learning phylogenetic trees from sequence alignment data

### 1. Loading alignment data

```{r}
library(phangorn)
library(ape)

ParisRT <- read.dna("ParisRT.txt")
```

### 2. Topology tree with three nucleotide substitution models

```{r}
pd_K80 <- dist.dna(ParisRT, model = "K80")
pd_TN93 <- dist.dna(ParisRT,model = "TN93")
pd_JC69 <- dist.dna(ParisRT,model = "JC69")

joined_K80 <- NJ(pd_K80)
joined_TN93 <- NJ(pd_TN93)
joined_JC69 <- NJ(pd_JC69)

# cex to decrease the text size
plot(joined_K80, cex=.7)
plot(joined_TN93, cex=.7)
plot(joined_JC69, cex=.7)
```

### 3. Log likelihood of Kimura nucleotide substitution model

```{r}
phyData_K80 <- as.phyDat(ParisRT)



pml_K80 <- pml(joined_K80, phyData_K80, model= "K80")

logLik(pml_K80)

print(pml_K80)
```

The log-likelihood of the fitted Kimura model is `r logLik(pml_K80)`.

### 4. Optimal parameters of Kimura nucleotide substitution model

```{r}
optim1_pml_K80 <- optim.pml(pml_K80, optQ =  TRUE, model = "K80")

print(optim1_pml_K80)

```

The values of the optimised rate matrix are:

| \-  |   a   |   c   |   g   |   t   |
|-----|:-----:|:-----:|:-----:|:-----:|
| a   | 0.000 | 1.000 | 5.098 | 1.000 |
| c   | 1.000 |   0   | 1.000 | 5.098 |
| g   | 5.098 | 1.000 |   0   | 1.000 |
| t   | 1.000 | 5.098 | 1.000 |   0   |

Different, compared to those of the unoptimised model:

| \-  |   a   |   c   |   g   |   t   |
|-----|:-----:|:-----:|:-----:|:-----:|
| a   | 0.000 | 1.000 | 1.000 | 1.000 |
| c   | 1.000 |   0   | 1.000 | 1.000 |
| g   | 1.000 | 1.000 |   0   | 1.000 |
| t   | 1.000 | 1.000 | 1.000 |   0   |

### 5. Optimisation of branch lengths, nucleotide substitution rates and tree topology; log likelihood of optimised model

```{r}
optim2_pml_K80 <- optim.pml(pml_K80, optNni = TRUE, optEdge = TRUE, optQ = TRUE, model = "K80")

print(optim2_pml_K80)
```

The log likelihood of the optimised model is `r logLik(optim2_pml_K80)`, higher, indicating a better fit, compared to the firstly optimised model, `r logLik(optim1_pml_K80)`, which is higher (better fit) than the starting unoptimised model `r logLik(pml_K80)`.

### 6. Bootstrapping

```{r}
bs_pml_K80 <- bootstrap.pml(optim2_pml_K80, optNni = TRUE, model = "K80", control = pml.control(trace = 0))

```

It is bootstrapping with replacement the columns of the alignment of the species, or in other words, the nucleotide sites of the alignment.

### 7. Plotting and guilty nurses

```{r}
plotBS(optim2_pml_K80$tree,bs_pml_K80,type = "phylogram",cex = 0.5)
```

The nurse "Mme S" is more likely to have effected "Mme L" because her HIV-virus strain sequence is phylogenetically more similar to that of patient "Mme L" compared to that of "Mr D" with "Mme L".
