---
title: "Homework 5: Statistical Phylogenetics"
format: pdf
editor: visual
---

---
title: "Homework 5: Statistical Phylogenetics"
author:
  - Bernhard Buss
  - Dario Franzen
  - Aurimas Greicius
  - Timothy Lauren Nijhuis
  - Carlo Rossetto
  - Yuliyan Slavchev
  - Görkem Kadir Solun
date: "March 2026"
format: pdf
---

## Problem 12: Intuition on time-reversible Markov chains

### 1. Time-reversed process transition probabilities

### 2. Original Markov chain reversability

\newpage

## Problem 13: Phylogenetic trees as Bayesian networks

### 1. Summation terms in brute force marginalisation

### 2. Summation terms after rearranging with Felsenstein's peeling algorithm.

\newpage

## Problem 14 (data analysis): Learning phylogenetic trees from sequence alignment data

### 1. Loading alignment data

```{r}
library(phangorn)
library(ape)

ParisRT <- read.dna("ParisRT.txt")
```

### 2. Topology tree with three nucleotide substitution models

```{r}
pd_K80 <- dist.dna(ParisRT, model = "K80")
pd_TN93 <- dist.dna(ParisRT,model = "TN93")
pd_JC69 <- dist.dna(ParisRT,model = "JC69")

joined_K80 <- NJ(pd_K80)
joined_TN93 <- NJ(pd_TN93)
joined_JC69 <- NJ(pd_JC69)

# cex to decrease the text size
plot(joined_K80, cex=.7)
plot(joined_TN93, cex=.7)
plot(joined_JC69, cex=.7)
```

### 3. Log likelihood of Kimura nucleotide substitution model

```{r}
phyData_K80 <- as.phyDat(ParisRT)



pml_K80 <- pml(joined_K80, phyData_K80, model= "K80")

logLik(pml_K80)

print(pml_K80)
```

The log-likelihood of the fitted Kimura model is `{r} logLik(pml_K80)`.

### 4. Optimal parameters of Kimura nucleotide substitution model

```{r}
optim1_pml_K80 <- optim.pml(pml_K80, optQ =  TRUE, model = "K80")

print(optim1_pml_K80)

```

The values of the optimised rate matrix (in lower triangular form, upper triangle is identical but mirrored) are: `{r}Q(optim_pml_K80)`. Compared to `{r} Q(pml_K80)` of the unoptimised model.

### 5. Optimisation of branch lengths, nucleotide substitution rates and tree topology; log likelihood of optimised model

```{r}
optim2_pml_K80 <- optim.pml(pml_K80, optNni = TRUE, optEdge = TRUE, optQ = TRUE, model = "K80")

print(optim2_pml_K80)
```

The log likelihood of the optimised model is `{r} logLik(optim2_pml_K80)`, higher, indicating a better fit, compared to the firstly optimised model, `{r} logLik(optim1_pml_K80)`, which is higher (better fit) than the starting unoptimised model `{r} logLik(pml_K80)`.

### 6. Bootstrapping

```{r}
bs_pml_K80 <- bootstrap.pml(optim2_pml_K80, optNni = TRUE, model = "K80")

print(bs_pml_K80)

```

It is bootstrapping with replacement the columns of the alignment of the species, or in other words, the nucleotide sites of the alignment.

### 7. Plotting and guilty nurses

```{r}
plotBS(optim2_pml_K80$tree,bs_pml_K80,type = "phylogram",cex = 0.5)
```
The nurse "Mme S" is more likely to have effected "Mme L" because her HIV-virus strain sequence is phylogenetically more similar to that of patient "Mme L" compared to that of "Mr D".
