---
title: "Homework 5: Statistical Phylogenetics"
author:
  - Bernhard Buss
  - Dario Franzen
  - Aurimas Greicius
  - Timothy Lauren Nijhuis
  - Carlo Rossetto
  - Yuliyan Slavchev
  - Görkem Kadir Solun
format: pdf
editor: visual
---

## Problem 12: Intuition on time-reversible Markov chains

### Part 1: Transition probabilities of the time-reversed process

Let $\{X(t), 0 \le t\}$ be a homogeneous continuous-time Markov chain starting from the stationary distribution, meaning $P(X(0) = k) = \pi_k$ for all states. Because the chain starts in the stationary distribution, the marginal probability of being in any state $k$ at *any* time $t$ is $\pi_k$.

We define the time-reversed process as $\hat{X}(t) := X(\tau - t)$. We want to find its transition probabilities: $$\hat{T}_{kl}(s,t) = Pr(\hat{X}(s+t) = l \mid \hat{X}(s) = k)$$

Substitute the definition of the reversed process: $$\hat{T}_{kl}(s,t) = Pr(X(\tau - (s+t)) = l \mid X(\tau - s) = k)$$

By the definition of conditional probability ($P(A|B) = \frac{P(A, B)}{P(B)}$): $$\hat{T}_{kl}(s,t) = \frac{Pr(X(\tau - s - t) = l, X(\tau - s) = k)}{Pr(X(\tau - s) = k)}$$

Let $u = \tau - s - t$. Notice that $u$ is the earlier time in the forward process, and $u+t = \tau - s$ is the later time. We rewrite the probabilities in chronological order for the forward process: $$\hat{T}_{kl}(s,t) = \frac{Pr(X(u) = l, X(u+t) = k)}{Pr(X(u+t) = k)}$$

Using the product rule ($P(A,B) = P(A)P(B|A)$), we can expand the numerator: $$\hat{T}_{kl}(s,t) = \frac{Pr(X(u) = l) \cdot Pr(X(u+t) = k \mid X(u) = l)}{Pr(X(u+t) = k)}$$

Because the process is stationary, $Pr(X(u) = l) = \pi_l$ and $Pr(X(u+t) = k) = \pi_k$. Furthermore, the forward conditional probability is just the standard transition matrix $T_{lk}(t)$: $$\hat{T}_{kl}(s,t) = \frac{\pi_l T_{lk}(t)}{\pi_k}$$

This result depends only on the duration $t$ and not on the specific starting time $s$. Therefore, the time-reversed process is also homogeneous, and we can write: $$\hat{T}_{kl}(t) = \frac{\pi_l T_{lk}(t)}{\pi_k} \quad \square$$

### Part 2: Equivalence for time-reversible chains

Assume the original Markov chain is time-reversible. By definition, this means: $$\pi_k T_{kl}(t) = \pi_l T_{lk}(t)$$

From Part 1, we established that the transition probability of the reversed process is: $$\hat{T}_{kl}(t) = \frac{\pi_l T_{lk}(t)}{\pi_k}$$

Rearranging the reversibility condition gives us $\frac{\pi_l T_{lk}(t)}{\pi_k} = T_{kl}(t)$. Substituting this into our equation for the reversed process yields: $$\hat{T}_{kl}(t) = T_{kl}(t)$$ This proves that a time-reversible Markov chain looks statistically identical whether viewed forwards or backwards in time. $\square$

## Problem 13: Phylogenetic trees as Bayesian networks

### Part 1: Joint probability and naive marginalization

Based on the directed edges of the Bayesian network $T$, the joint probability factorizes as the product of the probabilities of each node given its parent: $$P(Y, Z \mid T) = P(Z_1) \cdot P(Y_4 \mid Z_1) \cdot P(Z_2 \mid Z_1) \cdot P(Y_3 \mid Z_2) \cdot P(Z_3 \mid Z_2) \cdot P(Y_2 \mid Z_3) \cdot P(Y_1 \mid Z_3)$$

To find the marginal probability $P(Y \mid T)$, we must naively sum over all possible configurations of the hidden variables $Z = (Z_1, Z_2, Z_3)$: $$P(Y \mid T) = \sum_{Z_1} \sum_{Z_2} \sum_{Z_3} P(Z_1)P(Y_4|Z_1)P(Z_2|Z_1)P(Y_3|Z_2)P(Z_3|Z_2)P(Y_2|Z_3)P(Y_1|Z_3)$$

**Number of summation terms (naive):** There are 3 hidden variables ($Z_1, Z_2, Z_3$), and each can take one of 4 nucleotide states (A, C, G, T). To evaluate this naively, we must compute a term for every combination, requiring $4^3 = 64$ summation terms.

### Part 2: Felsenstein's Peeling Algorithm

We can use the distributive law to push the sums inward, evaluating them from the leaves (bottom) to the root (top). $$P(Y \mid T) = \sum_{Z_1} P(Z_1) P(Y_4 \mid Z_1) \left[ \sum_{Z_2} P(Z_2 \mid Z_1) P(Y_3 \mid Z_2) \left[ \sum_{Z_3} P(Z_3 \mid Z_2) P(Y_2 \mid Z_3) P(Y_1 \mid Z_3) \right] \right]$$

**Number of summation terms (peeling):**

**Innermost sum (over** $Z_3$): Computed for each of the 4 states of parent $Z_2$. Each sum iterates over the 4 states of $Z_3$. This yields $4 \times 4 = 16$ terms.

**Middle sum (over** $Z_2$): Computed for each of the 4 states of parent $Z_1$. Each sum iterates over the 4 states of $Z_2$. This yields $4 \times 4 = 16$ terms.

**Outermost sum (over** $Z_1$): Iterates over the 4 states of the root $Z_1$. This yields 4 terms.

At most, this algorithm requires $16 + 16 + 4 =$ $36$ summation terms.

------------------------------------------------------------------------

## Problem 14: Learning phylogenetic trees from sequence alignment data

```{r}
library(ape)
library(phangorn)
alignment <- read.dna("ParisRT.txt")
```

### Step 2: Initial Tree Topology (Neighbour Joining)

```{r}
# Calculate distances and build NJ trees for the three models
dist_k80  <- dist.dna(alignment, model = "K80")
tree_k80  <- NJ(dist_k80)

dist_tn93 <- dist.dna(alignment, model = "TN93")
tree_tn93 <- NJ(dist_tn93)

dist_jc69 <- dist.dna(alignment, model = "JC69")
tree_jc69 <- NJ(dist_jc69)

par(mfrow=c(1,3))
plot(tree_k80, main="NJ: Kimura (1980) (K80)")
plot(tree_tn93, main="NJ: Tamura and Nei 1993 (TN93)")
plot(tree_jc69, main="NJ: Jukes and Cantor 1969 (JC69)")
```

### Step 3: Fitting the Kimura Model (K80)

```{r}
# Convert data to phyDat format required by phangorn
phyDat_align <- phyDat(alignment)

# Fit the base pml model using K80
fit_k80 <- pml(tree_k80, phyDat_align, model = "K80")

# Print fit, incl. the loglik:
print(fit_k80)
```

### Step 4: Optimizing the Substitution Rate Matrix (Q)

```{r}
# Optimise ONLY the parameters of the substitution model (optQ)
fit_k80_optQ <- optim.pml(fit_k80, 
                          optQ = TRUE, 
                          optNni = FALSE, 
                          optBf = FALSE, 
                          optEdge = FALSE, 
                          optInv = FALSE, 
                          optGamma = FALSE, 
                          optRate = FALSE, 
                          optRooted = FALSE)

# Output optimized rate matrix
cat("Optimized K80 Rate Matrix (Transitions/Transversions):",
    print(fit_k80_optQ),"\n")
```

### Step 5: Full Optimization

```{r}
# Optimise topology (optNni), branch lengths (optEdge), and rates (optQ)
fit_full <- optim.pml(fit_k80, 
                        optQ = TRUE, 
                        optNni = TRUE, 
                        optBf = FALSE, 
                        optEdge = TRUE, 
                        optInv = FALSE, 
                        optGamma = FALSE, 
                        optRate = FALSE, 
                        optRooted = FALSE)

print(fit_full)
```

### Step 6 & 7: Bootstrapping and Forensic Analysis

**Question**: What exactly is being resampled?

**Answer**: The columns (sites) of the multiple sequence alignment are resampled with replacement.

```{r, results="hide"}
bs <- bootstrap.pml(fit_full, optNni = TRUE, bs = 100)
```

```{r fig.width=15, fig.height=10}
plotBS(fit_full$tree, 
       bs, 
       type = "phylogram", 
       main = "Optimised K80 Tree with Bootstrap"
       )
```

**Forensic Conclusion:** The nurse closest to the patient "**Mme_L**" is "**Mme S**", thus she is more likely to have infected the patient.
